voyAGEr: free web interface for the analysis of age-related gene expression alterations in human tissues

Version 1

Abstract

Abstract

We herein introduce voyAGEr, an online graphical interface to explore age-related gene expression alterations in 48 human tissues. voyAGEr offers a visualization and statistical toolkit for the finding and functional exploration of sex- and tissue-specific transcriptomic changes with age. In its conception, we developed a novel bioinformatics pipeline leveraging RNA sequencing data, from the GTEx project, for more than 700 individuals.

voyAGEr reveals transcriptomic signatures of the known asynchronous aging between tissues, allowing the observation of tissue-specific age-periods of major transcriptional changes, that likely reflect so-called digital aging, associated with alterations in different biological pathways, cellular composition, and disease conditions.

voyAGEr therefore supports researchers with no expertise in bioinformatics in elaborating, testing and refining their hypotheses on the molecular nature of human aging and its association with pathologies, thereby also assisting in the discovery of novel therapeutic targets. voyAGEr is freely available at https://compbio.imm.medicina.ulisboa.pt/voyAGEr

Introduction

The aging-associated progressive loss of proper tissue homeostasis maintenance makes age a prevalent risk factor for many human pathologies, including cancer, neurodegenerative and cardiovascular diseases 1–3. A better comprehension of the molecular mechanisms of human aging is thus required for the development and effective application of therapies targeting its associated morbidities.

Dynamic transcriptional alterations accompany most physiological processes occurring in human tissues 4. Transcriptomic analyses of tissue samples can thus provide snapshots of cellular states therein and insights into how their modifications over time impact tissue physiology. A small proportion of transcripts has indeed been shown to vary with age in tissue-5 and sex-specific 6–11 manners. Such variations reflect deregulations of gene expression that underlie cellular disfunctions 5.

Many studies analyzed the age-related changes in gene expression in rodent tissues 12–18, emphasizing the role in aging of genes related to inflammatory responses, cell cycle or the electron transport chain. However, while it is possible to monitor the modifications in gene expression over time in those species by sequencing transcriptomes of organs of littermates at different ages, as a close surrogate of longitudinality, such studies cannot be conducted in humans for obvious ethical reasons. Indeed, most studies aimed at profiling aging-related gene expression changes in human tissues focused on a single tissue (e.g. muscle 19–21, kidney 22, brain 23–27, skin 28,29, blood 30,31, liver 32, retina 33) and/or limited to a comparison between young and old individuals 20,21,28,32,33, failing to fully capture the changes of tissue-specific gene expression landscape along aging 5. A few studies were nonetheless led on more than one tissue in humans, from post-mortem samples 34,35 and biopsies 36,37, and in mice 18,36 and rats 38. The age-related transcriptional profiles derived therein, either from regression 18,34,35,37 or comparison between age groups 36,38, highlight an asynchronous aging of tissues (discussed in 39), with some of them more affected by age-related gene expression changes associated with biological mechanisms known to be impacted by aging such as mitochondrial activity or metabolic homeostasis. In particular, tissue-specific periods of major transcriptional changes in the fifth and eighth decades of human lifespan have been revealed 34, reflecting the so-called digital aging 39 and consistently with what is observed in mice 17,18. Furthermore, despite outlining the tissue specificity of the transcriptomic signatures of human aging, some synchronization was found between tissues like lung, heart and whole blood, which exhibit a co-aging pattern 35. Nevertheless, as each study followed its own specific procedures, from sample collection to data processing, results from their analyses are hardly comparable across studies.

Processed data from those studies have not been made easily accessible and interpretable to researchers lacking computational proficiency but aiming to use them to test their novel hypotheses. To fill this void, we have developed voyAGEr, a web application providing flexible visualization of comprehensive functional analyses of gene expression alterations occurring in 48 human tissues with age in each sex. We leverage the large RNA-seq dataset from the Genotype-Tissue Expression (GTEx) project 40, encompassing post-mortem tissue samples from hundreds of donors aged from 20 to 70 years, with a pipeline for gene expression profiling with an optimized temporal resolution. voyAGEr allows to investigate aging from two perspectives: (i) gene-centric – how each gene’s tissue-specific expression progresses with age; (ii) tissue-centric – how tissue-specific transcriptomes change with age. Additionally, voyAGEr enables the examination of modules of co-expressed genes altered with age in 4 tissues (brain, skeletal muscle, heart (left ventricle), whole blood), namely their enrichment in specific cell types, biological pathways, and association with diseases. We therefore expect voyAGEr to become a valuable support tool for researchers aiming to uncover the molecular mechanisms underlying human aging. Moreover, being open-source, voyAGEr can be adapted by fellow developers to be used with alternative datasets (e.g. from other species) or to incorporate other specific functionalities. voyAGEr is freely available at https://compbio.imm.medicina.ulisboa.pt/voyAGEr

Results

voyAGEr’s interactive exploration of tissue-specific gene expression landscapes in aging is based on the sequential fitting of linear models (v. Materials and Methods) to estimate, for each gene in each tissue (Figure 1):

voyAGEr profiles tissue-specific age-associated changes in gene expression and their differences between sexes.

For each of the 48 human tissues in GTEx, gene expression was linearly modelled in windows spanning 16 years centered in consecutive years of age, to estimate the effects thereon of Age, Sex and the interaction between them (Age&Sex), i.e., how the Sex effect changes with age, equivalent to how the Age effect differs between sexes (v. Materials and Methods). In each window (i.e., for each age in consecutive years), the percentage of genes with expression significantly altered by of each of those effects gives their respective transcriptomic impact (upper panels). voyAGEr thereby identifies the age periods at which major gene expression changes occur in each tissue. For example, in subcutaneous adipose tissue: major age-related transcriptional alterations are found at 30 and around 50 years of age (upper left panel), illustrated by the behavior of LMO3 (bottom left panel); major gene expression differences between males and females happen at around 60 years of age (upper center panel), as illustrated by GREB1 (bottom center panel); major differences between sexes in age-related gene expression alterations happen at around 50 years of age (upper right panel), as illustrated by ADAM33 (bottom right panel). Solid loess lines in the bottom panels (green for all donors, pink for females, blue for males). Gene expression (GE) in log2 of counts per million (logCPM).

the Age effect, i.e., how the age-associated changes in gene expression evolve with age itself;

the Sex effect, i.e., how the differences in gene expression between sexes evolve with age;

the Age&Sex interaction effect, i.e., how the Sex effect changes with age, equivalent to how the Age effect differs between sexes; in other words, how the age-associated changes in differences in gene expression between sexes (or how the differences between sexes of age-associated changes in gene expression) evolve with age.

We named our approach Shifting Age Range Pipeline for Linear Modelling (ShARP-LM). Briefly, this method carries out differential gene expression analyses (with gene expression as a function of the donors’ Age, Sex and Age&Sex interaction) in age windows spanning 16 years centered in consecutive years of age. By considering the percentage of genes differentially expressed in each age range, we highlight age periods of major tissue-specific transcriptomic alterations (Figure 1).

Gene-centric analyses of human tissue-specific expression changes across age

The progression of tissue-specific expression of a specific gene across age can be examined in votAGEr’s Gene tab. By entering its symbol in the Gene selector, the user has access to graphical summaries of the gene’s tissue-specific expression (sub-tab Profile) (Figure 2A) and of the significance age-related changes in its expression due to the Age, Sex, and Age&Sex (sub-tab Alteration) (Figure 2B) across age. Results can be displayed in a heatmap for all tissues or in a scatter plot for a chosen individual tissue. When the gene is studied in a single tissue, the user can graphically and statistically profile the association of the donors’ sex and reported conditions (e.g., history of heart attack or pneumonia) with the gene’s expression profile (Figure 2C). A table summarizing the donors’ metadata is also shown (Figure 2C). The user can interactively select donors of interest on the scatter plot and further examine their information in the automatically subsetted table.

Gene-centric analyses of expression alterations across age

A. Heatmap of MKI67 expression across tissues over age.

B. Heatmap of significance of Age-associated MKI67 expression alterations over age. p-values are for the moderated t-statistics of differential gene expression associated with the Age effect (v. the ShARP-LM approach in Materials and Methods). Notably, major transcriptional changes are observed in transverse colon (mid 20’s and after 55), breast (mid/late 40’s and early/mid 50’s), and liver (mid 50’s).

C. voyAGEr’s Gene tab interface. MKI67 expression in transverse colon is inspected. Donors’ information is shown in a table and the scatter plot can be interactively adjusted according to the donors’ condition of interest (Supplementary Figure 1B).

Cellular senescence, a stress-induced cell cycle arrest limiting proliferation of potentially oncogenic cells but progressively creating an inflammatory environment in tissues as they age, is an example of a process whose molecular mechanisms are of particular interest to aging researchers 41,42. CDKN2A, encoding cell cycle regulatory protein p16INK4A that accumulates in senescent cells 43,44, has its expression increased with age in the vast majority of tissues profiled (Supplementary Figure 1A). Similarly, reduced levels of proliferation markers, such as PCNA 45 and MKI67 46, can be studied as putative markers of aging of certain tissues. These genes have their expression significantly altered with age in transverse colon and display a similar expression profile (decreasing from 20 to 30 years old (y.o.), constant until 55 y.o. and decreasing in older ages) (Figure 2C and Supplementary Figure 1B). However, these trends appear to vary according to the donor’s history of pneumonia (Supplementary Figure 1B), illustrating voyAGEr’s use in helping to find associations between gene expression and age-related diseases.

On a different note, sex biases have been reported in for the expression of SALL1 and KAL1 in adipose tissue and lung, respectively 10. voyAGEr allows to not only recapitulate those observations but also estimate that these differences occur mostly in older ages (Supplementary Figure 2).

Tissue-specific assessment of gene expression changes across age

Peaks

The landscape of global tissue-specific gene expression alteration across age can be examined in voyAGEr’s Tissue main tab. A heatmap displaying, for all tissues, the statistical significance over age (v. Materials and Methods) of the proportion of genes altered with Age, Sex or Age&Sex interaction (depending on the user’s interest) is initially shown (Figure 3A), illustrating the aforementioned asynchronous aging of tissues observed for humans and rodents 18,32–36,39.

Tissue-specific assessment of gene expression changes across age

A. Heatmap of significance (FDR based on random permutations of age, v. Materials and Methods) over age of the proportion of genes with expression significantly altered with Age in the 48 analyzed tissues.

B. Exploration of gene expression changes across age in transverse colon.

a. Percentage of genes with significantly altered expression with Age over age. Three peaks at 27, 55 and 62 years old are noteworthy.

b. Clicking on the dot of a specific age (62.44 y.o. in plot a) gives access to the list of the most differentially expressed genes at that age, ordered by statistical significance of expression changes (p-value of moderated t-statistic).

c. Plot of expression of the chosen top gene in the table in b across age (bottom) in parallel with the significance of its expression alterations with Age. The expression of HOXA-AS3 significantly decreases at around 62 years old, concomitantly with the third peak of transcriptomic changes with Age.

The user can then spot the age periods with the most significant gene expression alterations in a selected tissue (Figure 3B a), the so-called Peaks naming the sub-tab, and identify the associated differentially expressed genes (Figure 3B b). The user can also plot the expression of a given gene of interest across age together with the significance of its expression modification with Age, Sex or Age&Sex (Figure 3B c).

As an example, the transverse colon appears to go through three main periods of transcriptional changes: one minor at around 27 y.o. (∼5 % of genes differentially expressed (DEG)) and two major at around 55 y.o. (∼21 % DEG) and 62 y.o. (∼40 % DEG) (Figure 3B a). Interestingly, some of the genes most differentially expressed in the second peak at ∼55 y.o. appear to have their expression modified only at this precise age period (e.g. E2F1, KIF24, TRMU, CRIM1). Similarly, genes such as HOXA-AS3, NFYA, ZDHHC1, and MYL6B (Figure 3B b), are specifically differentially expressed in the third peak at ∼62 y.o. (Figure 3B c). This particularity is also found for the minor first peak. It therefore seems that different sets of genes drive the periods of major transcriptional changes, which begs assessing if they reflect the activation of distinct biological processes.

Enrichment

The user can thus explore the biological functions of the set of genes underlying each peak of transcriptomic changes by assessing their enrichment in manually curated pathways from the Reactome database 47. voyAGEr performs Gene Set Enrichment Analysis (GSEA) 48 and the user can visualize heatmaps displaying the evolution over age of the resulting normalized enrichment score (NES, reflecting the degree to which a pathway is over- or under-represented in a subset of genes) for the given tissue, effect (Age, Sex or Age&Sex) and all or some user-specified Reactome pathways (Figure 4A). To reduce redundancy and facilitate the understanding of their biological relevance, we clustered those pathways into families that also include Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways 49 and Gene Ontology (GO) Biological Processes of level 3 50. Briefly, we clustered gene sets from the 3 sources based on the overlap of their genes (v. Materials and Methods), thereby creating families of highly functionally related pathways. Taking advantage of the complementary and distinct terminology in Reactome, KEGG and GO, the user can interpret each family’s broad biological function by looking at the word cloud of its most prevalent terms, still being able to browse the list of its associated pathways (Figure 4A). The peaks of transcriptomic changes can also be examined for enrichment in a user-provided gene set (Figure 4B). To illustrate this procedure, we inspected the aforementioned three aging peaks of the transverse colon for significant enrichment in genes retrieved from Senequest 42 whose link with senescence is supported by at least 4 sources, having observed it in the first (∼27 y.o.) and second (∼55 y.o.) peaks (Figure 4B).

Tissue-specific assessment of pathway expression changes across age

A. Heatmap of enrichment (above), given by GSEA’s normalized enrichment score (NES), of Reactome pathways in genes differentially expressed with Age over age. Pathways are classified into broad families (upper bar), whose associated biological processes and functions can be learnt from the word clouds of their most frequent terms (below). Only Reactome pathways found significantly associated (FDR ≤ 0.01) with gene expression alterations in at least one age window are included.

B. Enrichment, given by the significance of Fisher’s tests, of a user-provided gene set in genes differentially expressed (based on a user-defined p-value threshold) with Age over age. Here, the given gene set is composed of genes from Senequest 42 whose link with senescence is supported by at least 4 sources. The first and second peaks of Age-associated transcriptomic changes in the transverse colon appear to be enriched in senescence-related genes.

Modules of co-expressed genes

voyAGEr also allows functional analyses of modules of co-expressed genes (i.e. genes with highly correlated expression across samples, defined by weighted correlation network analysis 51) altered with age. Genes in the same module are likely to be coregulated and share biological functions or associations with phenotypical or pathological traits 52. Those modules may also act as markers of core transcriptional features of cellular activity and identity 53.

Concretely, voyAGEr enables the user to visualize how the expression of modules of genes that are associated with a specific cell type or disease progresses over age in a specific tissue. In the example of Figure 5B, the “midnight blue” module, gathering 11 genes co-expressed in the heart’s left ventricle, shows significant enrichment in fibroblast markers, namely collagens. This module’s expression in the left ventricle is steady until ∼55 years of age and appears to increase later in life, perhaps reflecting the known age-related changes of the collagen network of the human myocardium 54.

Tissue-specific assessment of age-associated progression of modules of co-expressed genes

A. Heatmap of eigengene expression for all the inferred modules of co-expressed genes in the heart’s left ventricle over age.

B. Scatter plot (above) of eigengene expression over age, in all left ventricle samples, for a selected module of 11 genes. The eigengene expression is derived from the 1st component of the single value decomposition of its genes’ expression. This module was analysed, with Fisher’s tests, for enrichment in cell types, based on markers from the literature (Skelly 84, Cui 83 and He 82, surnames of the first authors of the studies the markers are retrieved from) and found to be associated with fibroblasts, as revealed by the TreeMap below (where each rectangle’s area and darkness are proportional to the significance of its association with a cell type and its color linked to the markers’ source study).

C. Heatmap of association of the modules with five selected diseases, computed with the DOSE package 57.

After selecting one tissue of interest, the user has, for each module, access to four levels of information: (i) Expression: its eigengene expression progression over age; (ii) Cell types: its enrichment in specific cell types, based on marker genes found literature, (iii) Pathways: its enrichment in Reactome pathways; (iv) Diseases: its enrichment in disease markers, based on gene-disease associations from DisGeNET 55,56, calculated with both the DOSE package 57 and with Fisher’s tests (Figure 5C).

By default, for each tissue, results are displayed in the form of heatmaps showing how all modules evolve with age (Figure 5A). The user has also the possibility to select a module of interest and see its eigengene progression over age in a scatter plot (Figure 5B), lists of its association with diseases and pathways ordered by significance, and a TreeMap for its cell type enrichment (Figure 5B). Like in the Gene tab, the user can separate donors based on their sex and previous condition in the scatter plot of the eigengene expression progression. When a significant enrichment results from a Fisher’s test, the associated contingency table is shown for its clearer interpretation.

Discussion

voyAGEr provides a framework to examine the progression over age of gene expression in a broad range of human tissues, thereby being a novel valuable resource for the aging research community. In particular, it helps to identify tissue-specific age periods of major transcriptomic alterations. The results of our analyses show the complexity of human biological aging by stressing its tissue specificity 34 and non-linear transcriptional progression throughout lifetime, consistent with previous results from both proteomic 58 and transcriptomic 28,34 analyses. By revealing and annotating the age-specific transcriptional trends in each tissue, voyAGEr aims at assisting researchers in deciphering the cellular and molecular mechanisms associated with the age-related physiological decline across the human body.

Besides, voyAGEr, to our knowledge unprecedently, scrutinizes and visually displays the tissue-specific differences in gene expression between biological sexes across age. Biological sex is an important variable in the prevalence of aging-associated diseases, as well as in their age of onset and progression 6,7,59 and tissue-specific sex-related biases in gene expression had been reported 8–11. By profiling the age distribution of those biases, voyAGEr will thus help to better understand their influence in the etiology of the sex specificities of human aging. For instance, we were able to corroborate findings on the sex-differential transcriptome of human adults by Gershoni et al. 9, with voyAGEr emphasizing its tissue-specificity and allowing to discriminate the ages at which sex-related biases appear to be more prevalent (Supplementary Figure 3).

One of the limitations of voyAGEr is that most GTEx tissue donors had health conditions and their frequency increased with age, preventing us from defining a class of healthy individuals and identifying age-associated transcriptomic changes that could be more confidently proposed to happen independently of any disease progression. On one hand, we expect the large sample sizes and the biological variability between individuals, reflected on the diversity of combinations of conditions, to mitigate major confounding effects. On the other hand, voyAGEr allows users to assess how tissue-specific gene expression trends vary according to the donors’ history of the different conditions (Supplementary Figure 1B), therefore helping to find associations between gene expression and age-related diseases.

The development of voyAGEr was accompanied by that of a pipeline, ShARP-LM, that facilitates the holistic depiction of the transcriptional landscapes of adult human tissues upon aging with a yearly age resolution. We take advantage of the comprehensiveness of the collection of transcriptomes from post-mortem human tissues from the GTEx project to make our analyses a valid surrogate of a currently undoable longitudinal study. It confers our method enough statistical robustness to mitigate the inter-individual, overcome the imbalance in the numbers of male and female samples and deal with the non-uniform distribution of the donors’ ages. Indeed, we found no confounding between the distribution of samples’ ages and the trend of gene expression progression over age in any tissue.

However, the distribution of donor’s ages does affect the statistical power of detection of transcriptional alterations and we are therefore more likely to find significant ones in age ranges that are more abundant in samples (Supplementary Figure 4). This limitation is to be overcome by the accumulation of transcriptomes of human tissues in public databases, promising a gradually increase in accuracy and age resolution with which human transcriptomic aging can be profiled. Similarly, the accumulation of single-cell transcriptomes in public databases is yielding better gene markers for a growing diversity of human cell types, increasing the usefulness of exploiting bulk transcriptomes in studying the impact of aging on the cellular composition of human tissues, for which the co-expression module approach in voyAGEr provides a proof-of-concept.

As voyAGEr is open source, it is envisaged as an ever-evolving resource, that can be adapted by researchers to accommodate the aforementioned new data and expanded to provide new functionalities (e.g. integration with perturbagen data for the inference of molecular causes of observed gene expression alterations or small molecules for therapeutic purposes 60,61).

As an in silico approach with no experimental validation for its results, voyAGEr is meant to be a discovery tool, supporting biologists of aging in the exploration of a large transcriptomic dataset, thereby generating, refining or preliminary testing hypotheses that will pave the way for subsequent experimental work. It can be an entry point for projects aiming at better understanding the tissue- and sex-specific transcriptional alterations underlying human aging, to be followed by more targeted studies focusing on the functional roles of the most promising markers identified therein in the physiology of aging. Those marker genes can contribute to the development of more robust and cross-tissue gene signatures of aging 62 and the expansion of age-related gene databases 57,63.

Moreover, the observed diverse and asynchronous changes in gene expression between tissues over the human adult life provides potentially relevant information for the design of accurate diagnostic tools and personalized therapies. On one hand, finding those changes’ association with specific disorders could have a prognostic value by enabling the identification of their onset before their clinical symptoms’ manifestation 36. On the other hand, computational screening of databases of genetic and pharmacologically induced human transcriptomic changes could help to infer their molecular causes and shed light on candidate drugs to delay their effects 60,61,64,65.

Methods

Development’s platform

Data analysis was performed in R (version 3.6.3) and the application developed with R package Shiny 66. voyAGEr’s outputs are plots and tables, generated with R packages highcharter 67 and DT 68, respectively, that can easily be downloaded in standard formats (png, jpeg, and pdf for the plots; xls and csv for the tables).

voyAGEr was deployed using Docker Compose and ShinyProxy 2.6.0 in a Linux virtual machine (64GB RAM, 16 CPU threads and 200GB SSD) running in our institutional computing cluster. The source code for voyAGEr is available at GitHub (https://github.com/nunolbmorais/voyAGEr).

Read count data pre-processing

We downloaded the matrix with the RNA-seq read counts for each gene in each GTEx v7 sample from the project’s data portal (https://www.gtexportal.org/) 40. We then pre-processed the read count data for each tissue separately. We started by filtering out the uninformative genes with very low expression across samples. For that purpose, we normalised read counts for library size, to account for the differences in sequencing depth between samples, by transforming them into counts per million (CPM) with function cpm from R package edgeR 69. Only genes with at least 1 CPM in at least 40% of the samples were kept for analysis (the number of genes analysed for each tissue can be found in Supplementary Table 1). Read counts for those kept genes were used to calculate normalization factors to scale the raw library sizes, using function calcNormFactors from edgeR that implements the trimmed mean of M-values 70, and normalised accordingly and subsequently log-transformed with the voom function 71 from package limma 72, which also estimates mean-variance relationships to be used in linear modelling (v. ShARP-LM below). Normalized gene expression was thereby ready to be analyzed and featured in the plots in log2 of CPM (logCPM), this log-transformation yielding more manageable nearly normal distributions.

Finally, from the 53 tissues available from GTEx v7, we discarded five (renal cortex, fallopian tube, bladder, ectocervix, endocervix, chronic myeloid leukemia cell line) due to a low (<50) number of respective samples.

ShARP-LM

To model the changes in gene expression with age, we developed the Shifting Age Range Pipeline for Linear Modelling (ShARP-LM). For each tissue, we fitted linear models to the gene expression of samples from donors with ages within windows with a range of 16 years shifted through consecutive years of age (i.e. in a sliding window with window size = 16 and step size = 1 years of age). As samples at the ends of the dataset’s age range would be thereby involved in fewer linear models, we made the window size gradually increase from 11 to 16 years when starting from the “youngest” samples and decrease from 16 to 11 years when reaching the “oldest” (Supplementary Figure 5).

Function lm from limma was used to fit the following linear model for gene expression (GE): For each gene, α, β and χ are the coefficients to be estimated for the respective hypothesized effects and ε is the residual. For each sample, Age in years and Sex in binary were centered and Age&Sex interaction given by their product.

For each gene in each model (i.e., each age window in each tissue), we retrieved the t-statistics of differential expression associated with the three variables and their respective p-values. We considered the average age of the samples’ donors within the age window as the representative age of the observed expression changes.

In summary, for a given tissue and variable (Age, Sex and Age&Sex), ShARP-LM yields t-statistics and p-values over age for all genes, reflecting the magnitude and significance of the changes in their expression throughout adult life.

Gene-centric visualization of tissue-specific expression changes across age

For visualization purposes, the trend of each gene’s expression progression over age in each tissue was derived through Local Polynomial Regression Fitting, using R function loess on logCPM values (Figures 1, 2C, 3B). For summarizing in a heatmap a given gene’s expression across age in multiple tissues (Figure 2A) or the expression of several genes across age in a given tissue, each gene’s regression values are centered, but not scaled, with R function scale.

For summarizing in a heatmap the significance of a given gene’s expression changes over age in multiple tissues (Figure 2B), cubic smoothing splines are fitted to -log10(p), p being the t-statistic’s p-value, with R function smooth.spline.

Tissue-specific quantification of global transcriptomic alterations across age

To assess the global transcriptomic impact of each of the three modelled effects in each tissue across age, we then analyzed the progression over age (i.e. over consecutive linear models) of the percentage of genes whose expression is significantly altered (t-statistic’s p-value ≤ 0.01) by each effect (Figure 3B). To evaluate the significance of each percentage and see if high ones can be confidently associated with major transcriptomic alterations, we controlled for their false discovery rate (FDR) by randomly permuting the samples’ ages within each age window one million times and performing ShARP-LM on each randomised dataset. We were then able to associate an FDR to each percentage of differentially expressed genes by comparing it with the distribution of those randomly generated (Figure 3B).

Gene Set Enrichment Analysis (GSEA)

For each Peak of significant gene expression modifications, we performed GSEA 48 on the ordered (from the most positive to the most negative) t-statistics of differential expression for the respective tissue and age, using R package fgsea 73 and the Reactome database 47. The resulting normalized enrichment score (NES) for each pathway was used in subsequent analyses as a metric of its over- or down-representation in the Peak.

Families of pathways

To reduce pathways’ redundancy and facilitate the assessment of their biological relevance in results’ interpretation, we created an unifying representation of pathways from Reactome 47 and KEGG 49, and level 3 Biological Processes from GO 50, by adapting a published pathway clustering approach 74 to integrate them into families.

The approach relies on the definition of a hierarchy of pathways based on the numbers of genes they have in common. For each two pathways Pi and Pj, respectively containing sets of genes Gi and Gj, we computed their overlap index 74, defined as following: Where |Gi| is the number of genes in set Gi and |Gi⋂Gj| is the number of genes in common between Gi and Gj. Oiij = 1 would therefore indicate that Pi and Pj are identical in gene composition or that one is a subset of the other. On the contrary, Oiij = 0 would mean that Pi and Pj have no genes in common. To ease the computational work, we removed from the analysis pathways that are subsets of larger pathways (i.e. each pathway whose genes are all present in another pathway).

From the Oiij matrix, from which each row is a vector with the gene overlaps of pathway I with each of all the pathways, we computed the matrix of Pearson’s correlation between all pathways’ overlap indexes with R function cor. That matrix was finally transformed into Euclidean distances with R function dist, allowing for pathways to be subsequently clustered with the complete linkage method with R function hclust. The final dendrogram was empirically cut in 10 clusters (Supplementary Figure 6). One of them, not including Reactome pathways, was dismissed from the analysis. Pathways that initially excluded from the computation for being subsets of others were finally added to the clusters of their respective parent pathways. Each daughter pathway with more than one parent was assigned to the cluster of the parent with the smallest number of genes, thereby maximizing the daughter-parent similarity. The data.table R package, for fast handling of large matrices, was used in this analysis 75.

Gene co-expression modules

Gene co-expression modules were defined with R package WGCNA 51. For each tissue, genes were first filtered based on the variance of their expression across samples (variable A), with lowly variant ones being discarded from subsequent co-expression analyses. Then, an adjacency matrix was computed with the adjacency function. To this end, we first computed the matrix of Pearson’s correlation between all selected genes’ expressions and then raised all of its values to a power β (soft thresholding) to emphasise stronger correlations. The soft threshold was determined with the pickSoftThreshold function. The thereby generated matrix (|correlation|β) was then converted into a dissimilarity matrix based on the topological overlap measure, determined with function TOMsimilarity. Finally, gene co-expression modules were identified with a dynamic tree-cutting algorithm, implemented with the cutreeDynamic function, that adjusts the minimal number of genes per module (variable B) and the height at which the tree is cut (variable C).

The definition of those modules of co-expressed genes is highly dependent on the three variables A, B, and C. Thus, we comprehensively tested combinations of them and defined as optimal those that maximized the modules’ enrichment in known markers for specific cell types, an approach similar to that reported to have been successfully applied in the revelation of the core transcriptional features of cell types in the human central nervous system 53. In particular, we used human and murine cell type gene signatures, derived from single cell transcriptomic analyses, for brain (cortex) 76–79, skeletal muscle 80–82, left cardiac ventricle 82–84, and whole blood 82,85,86. Finally, the 40%, 30%, 35%, and 40% most variant genes, the tree cut at a height of 0.96, and a minimum number of 20, 18, 11, and 16, were respectively chosen for the three variables in those four tissues.

Each module is then represented by a set of genes and an eigengene, the first principal component obtained by singular-value decomposition of the expression of the module’s genes and representative of its own expression profile.

The enrichment of modules in cell types, Reactome pathways and diseases from DisGeNET 55,56was quantified with Fisher’s tests. Disease enrichment was also calculated using the DOSE package 57.

eLife Assessment

This work presents a valuable online platform designed to facilitate the exploration of genes and genetic pathways implicated in human aging. Leveraging a new inference methodology, the tool enables the identification and visualization of key genes and tissues impacted by aging, facilitating scientific discovery. However, since the methods have not been fully explained, there is only incomplete support for the platform, which could be improved upon for usability and reproducibility.

Reviewer #1 (Public Review):

This fascinating paper by A.L. Schneider et al. describes voyAGEr, a shiny-based interface for easy exploration of the GTEx dataset by non- or novice programmers. Importantly, voyAGEr is open source and available from github, which could greatly accelerate additional development and further uses of this interesting tool.

The authors developed a pipeline for modeling age-related changes in gene expression in the GTEx data called ShARP-LM, fitting a linear model for age, sex, and age&sex interaction terms. This pipeline underlies the later analyses that can be applied within voyAGEr. These analyses are labeled by tissue so that users can easily begin a query based on a tissue or a gene of possible interest.

voyAGEr implements many kinds of interesting R-based tools such as pathway overrepresentation analysis and gene co-expression module analysis, in a way that makes these approaches accessible to non-bioinformaticist aging researchers.

As the tidal wave of publicly available large, high-dimensional datasets such as transcriptomes continues to grow exponentially, the usefulness of tools such as voyAGEr will only increase. While test users may be able to imagine features or refinements they wish were already present, due to the open source approach they or anyone else including but not limited to the present authors can implement additional features in the future. I look forward to using this tool and to staying abreast of its future development.

Overall, this study describes a new tool of interest to the field. The manuscript is clearly written overall. The figures and supplementary information are all clear and all add to the manuscript.

Reviewer #2 (Public Review):

The purpose of this study is to develop a tool that serves as a starting point for investigating and uncovering genes and pathways associated with aging. The tool utilizes information from the GTEx public database, which contains post-mortem human data. It focuses on identifying age-related gene expression changes across different age range, biological sexes, and medical histories, with a focus on specific tissues.

Additionally, the authors envision the platform as continuously evolving, with ongoing development and expansion to include new data and features, ensuring it remains a cutting-edge resource for researchers studying aging.

# Strengths

voyAGEr presents a tool for exploring gene expression changes across multiple tissues in the context of aging. One of the main strengths of the tool is its intuitive and user-friendly interface, which allows for easy navigation and exploration of gene expression patterns for biologists. Users can explore changes in gene expression of single genes across multiple tissues, enabling them to identify genes of interest that can be further investigated.

A particularly noteworthy strength of the tool is its ability to show tissue-specific gene expression patterns. This feature is essential for elucidating the paradigm of tissue-specific asynchronous aging and provides a unique and valuable resource for the aging community.

Overall, the tool offers an entry point for further investigation of genes involved in aging, and its ability to show tissue-specific gene expression patterns provides a unique and valuable resource for the scientific community.

Lastly, the tool is accompanied by a clear and thorough tutorial that explains each of its functionalities and provides examples. The authors also acknowledge the limitations of the statistical inference tests used in the tool, which adds to its overall transparency.

# Weaknesses

## Underlying data analysis

In this tool/resource paper, it is crucial that the data used is up-to-date to provide the most comprehensive and relevant information to users. However, the authors utilized GTEx v7, which is an outdated (2016) version of the dataset. It is worth noting that GTEx v8 includes over 940 individuals, representing a 35% increase in individuals, and a 50% increase in the total number of samples. The authors should check the newer versions of GTEx and update the data.

The authors did not address any correction for batch effects or RNA integrity numbers, which are known to affect transcriptome profiles. For instance, our analysis of GTEx v8 Cortex tissue revealed that after filtering out lowly expressed genes, in the same way authors did, PC1 (which accounts for 24% of the variation) had a Spearman's correlation value of 0.48 (p<6.1e-16) with RNA integrity number.

The data analyzed in the GTEx dataset is not filtered or corrected for the cause of death, which can range from violent and sudden deaths to slow deaths or cases requiring a ventilator. As a result, the data may not accurately represent healthy aging profiles but rather reflect changes in the transcriptome specific to certain diseases due to the age-related increase in disease risk. While the authors do acknowledge this limitation in the discussion, stating that it is not a healthy cohort and disease-specific analysis is not feasible due to the limited number of samples, it would be useful for users to have the option to analyze only cases of fast death, excluding ventilator cases and deaths due to disease. This is typically how GTEx data is utilized in aging studies. Alternatively, the authors should consider including the "cause of death" variable in the model.

The age distribution varies across tissues which may impact the results of the study. The authors' claim that age distribution does not affect the outcomes is inconclusive. Since the study aims to provide cross-tissue analysis, it is important to note that differing age distributions across tissues can influence the overall results. To address this, the authors should conduct downsampling to different age distributions across tissues and evaluate the level of tissue-specific or common changes that remain after the distributions are made similar.

The GTEx resource is extremely valuable, however, it comes with challenges. GTEx contains tissue samples from the same individuals across different tissues, resulting in varying degrees of overlap in sample origin across tissues as not all tissues are collected for all individuals. This could affect the similar/different patterns observed across tissues. As this tool is meant for broader use by the community, it is crucial for the authors to either rule out this possibility by conducting a cross-tissue comparison using a non-parametric model that accounts for the dependency between samples from the same individual, or to provide information on the degree of similarity between samples so that the users can keep this possibility in mind when using the tool for hypothesis generation.

## Visualisation and analysis platform

The authors aimed to create an open-source and ever-evolving resource that could be adapted and improved with new functionality. However, this goal was only partially achieved. Although the code for the web app is open source, crucial components such as the statistical tests or the linear model are not included in the repository, limiting the tool's customizability and adaptability.

Furthermore, the authors' choice of visualization platform (R shiny) may not be the best fit for extensibility and open-source collaboration, as it lacks modularity. A more suitable alternative could be production-oriented platforms such as Flask or FastAPI.

To facilitate collaboration and improve the tool's adaptability, data resulting from the pre-processing pipeline should be made publicly available. This would make it easier for others to contribute and extend the tool's functionality, ultimately enhancing its value for the scientific community.

Reviewer #3 (Public Review):

In their manuscript, Schneider et al. aim to develop voyAGEr, a web-based tool that enables the exploration of gene expression changes over age in a tissue- and sex-specific manner. The authors achieved this goal by calculating the significance of gene expression alterations within a sliding window, using their unique algorithm, Shifting Age Range Pipeline for Linear Modelling (ShARP-LM), as well as tissue-level summaries that calculated the significance of the proportion of differentially expressed genes by the windows and calculated enrichments of pathways for showing biological relevance. Furthermore, the authors examined the enrichment of cell types, pathways, and diseases by defining the co-expressed gene modules in four selected tissues. The voyAGEr was developed as a discovery tool, providing researchers with easy access to the vast amount of transcriptome data from the GTEx project. Overall, the research design is unique and well-performed, with interesting results that provide useful resources for the field of human genetics of aging. I have a few questions and comments, which I hope the authors can address.

1. In the gene-centric analyses section of the result, to improve this manuscript and database, linear regression tests accounting for the entire range of age should be added. The authors' algorithm, ShARP-LM, tests locally within a 16-year window which makes it has lower power than the linear regression test with the whole ages. I suspect that the power reduction is strongly affected in the younger age range since a larger number of GTEx donors are enriched in old age. By adding the results from the lm tests, readers would gain more insight and evidence into how significantly their interest genes change with age.

2. In line with the ShARP-LM test results, it is not clear which criterion was used to define the significant genes and the following enrichment analyses. I assume that the criterion is P < 0.05, but it should be clearly noted. Additionally, the authors should apply adjusted p-values for multiple-test correction. The ideal criterion is an adjusted P < 0.05. However, if none or only a handful of genes were found to be significant, the authors could relax the criteria, such as using a regular P < 0.01 or 0.05.

3. In the gene-centric analyses section, authors should provide a full list of donor conditions and a summary table of conditions as supplementary.

4. The tissue-specific assessment section has poor sub-titles. Every title has to contain information.

5. I have an issue understanding the meaning of NES from GSEA in the tissue-specific assessment section. The authors performed GSEA for the DEGs against the background genes ordered by t-statistics (from positive to negative) calculated from the linear model. I understand the p-value was two-tailed, which means that both positive and negative NES are meaningful as they represent up-regulated expression direction (positive coefficient) and down-regulated expression direction (negative coefficient) with age, respectively, within a window. However, in the GSEA section of Methods, authors were not fully elaborate on this directionality but stated, "The NES for each pathway was used in subsequent analyses as a metric of its over- or down-representation in the Peak". The authors should clearly elaborate on how to interpret the NES from their results.

6. In the Modules of co-expressed genes section, the authors did not explain how or why they selected the four tissues: brain, skeletal muscle, heart (left ventricle), and whole blood. This should be elaborated on.

7. In the modules of the co-expressed genes section, the authors did not provide an explanation of the "diseases-manual" sub-tab of the "Pathway" tab of the voyAGEr tool. It would be helpful for readers to understand how the candidate disease list was prepared and what the results represent.

Author Response

The following is the authors’ response to the original reviews.

We are pleased to send you a revised version of our manuscript entitled “voyAGEr: free web interface for the analysis of age-related gene expression alterations in human tissues” and the associated shiny web app, in which we incorporate the referees’ feedback. We would like to express our gratitude for their time and valuable insights, which have contributed to the improvement of our work. We appreciate the rigorous evaluation process that eLife maintains.

In this letter, we address each of the reviewers' comments and concerns, point-by-point, offering detailed responses and clarifications. We have made several revisions to our manuscript following their recommendations.

We must note that the revised version of the manuscript has two novel joint first authors, Rita Martins-Silva and Alexandre Kaizeler, who performed all the requested reanalyses, given that the initial first author, Arthur Schneider, already left our lab. We must also point to the following minor unsolicited improvements we took the opportunity to make:

Added a comprehensive tutorial to the GitHub repository on how to navigate through voyAGEr’s features.

Implemented sample randomisation in the scatter plots depicting gene expression across the age axis to ensure data privacy.

Implemented minor adjustments within the web app to enhance user comprehension and clarity when visualizing the data.

Improved clarity of the methodological sections.

Reviewer 1

(1.1) While this may be obvious to others for some reason that escaped me, I was unsure what was the basis for the authors' choice of 16 years as the very specific sliding window size. If I'm not alone in this, it might add clarity for other readers and users if this parameter choice were explained and justified more explicitly.

We apologise for our omission in providing the rationale behind our choice in the previous version. We chose 16 years as our sliding window size because this was the minimum needed to guarantee the presence of more than one sample per window, across all the tissues considered in the study (Figure R1 below).

We added the following sentence to the manuscript (v. Methods, ShARP-LM):

“This was the minimum age span needed to guarantee the presence of more than one sample per window, across all considered tissues.”

(1.2) "In particular, tissue-specific periods of major transcriptional changes in the fifth and eighth decades of human lifespan have been revealed, reflecting the so-called digital aging and consistently with what is observed in mice" here I think that "consistently" should be "consistent".

We thank the reviewer for the comment and following the suggestion, we have revised 'Consistently' to 'consistent' as it is the correct usage in our sentence.

(1.3) "On a different note, sex biases have been reported in for the expression of SALL1 and KAL1 in adipose tissue and lung, respectively." Here I think that "in for" should be "in".

As recommended by the reviewer, we have replaced ‘in for’ for ‘in’. As we substituted KAL1, the current sentence now stands as “On a different note, sex biases have been reported in the expression of SALL1 and DDX43 in adipose tissue and lung, respectively”.

(1.4) "We downloaded the matrix with the RNA-seq read counts for each gene in each GTEx v7 sample from the project's data portal (https://www.gtexportal.org/)." In my pdf manuscript this hyperlink appears to be broken.

We appreciate the reviewer's attention to the broken link, and we have rectified the issue. The link should now be fully operational, effectively directing users to the GTEx Portal.

(1.5) Under methods, I might suggest "Development platform" or "Development platforms" over "Development's platform" as a heading.

We have modified the heading of this section in the methods to 'Development Platforms', as we believe it better reflects the information conveyed.

Reviewer 2

(2.1) In this tool/resource paper, it is crucial that the data used is up-to-date to provide the most comprehensive and relevant information to users. However, the authors utilized GTEx v7, which is an outdated (2016) version of the dataset. It is worth noting that GTEx v8 includes over 940 individuals, representing a 35% increase in individuals, and a 50% increase in the total number of samples. The authors should check the newer versions of GTEx and update the data.

When the development of the voyAGEr web application began, GTEx version 7 was the most up to date. Nevertheless, we agree that the version 8 offers a notably more extensive dataset, encompassing a larger number of individuals, samples, and introducing new tissues. Consequently, we have updated our application to incorporate the data from GTEx version 8.

(2.2) The authors did not address any correction for batch effects or RNA integrity numbers, which are known to affect transcriptome profiles. For instance, our analysis of GTEx v8 Cortex tissue revealed that after filtering out lowly expressed genes, in the same way authors did, PC1 (which accounts for 24% of the variation) had a Spearman's correlation value of 0.48 (p<6.1e-16) with RNA integrity number.

We acknowledge the validity of the reviewer’s comment and appreciate the importance of such corrections to enhancing data interpretation. In response, we conducted a thorough unbiased investigation into potential batch effects, with the COHORT variable emerging as the primary driver of those observed across most tissues. Furthermore, SMRIN (as the reviewer pointed), DTHHRDY, MHSMKYRS and the number of detected genes in each sample were consistently associated with the primary sources of variation. As a result, we implemented batch effect correction for those five conditions, in a tissue-specific manner.

We provide a detailed explanation of the batch effect correction methodology and its importance in the biological interpretation of results in the Methods section, specifically under "Read count data pre-processing". Additionally, we have included two new supplementary figures, Sup. Figures 7 and 8, to illustrate a batch effect example in lung tissue and emphasise the critical role of this correction in data interpretation.

(2.3) The data analyzed in the GTEx dataset is not filtered or corrected for the cause of death, which can range from violent and sudden deaths to slow deaths or cases requiring a ventilator. As a result, the data may not accurately represent healthy aging profiles but rather reflect changes in the transcriptome specific to certain diseases due to the age-related increase in disease risk. While the authors do acknowledge this limitation in the discussion, stating that it is not a healthy cohort and disease-specific analysis is not feasible due to the limited number of samples, it would be useful for users to have the option to analyze only cases of fast death, excluding ventilator cases and deaths due to disease. This is typically how GTEx data is utilized in aging studies. Alternatively, the authors should consider including the "cause of death" variable in the model.

This comment is closely related to the prior discussion (point 2.2). Notably, two of the covariates selected for batch effect correction, namely, DTHHRDY (Death classification based on the 4-point Hardy Scale1) and COHORT (indicating whether the participant was a postmortem, organ, or surgical donor1), have a direct relevance to this issue, i.e., both relate to the cause of death of the individual.

1 According to the nomenclature of variables described in https://www.ncbi.nlm.nih.gov/projects/gap/cgibin/ GetListOfAllObjects.cgi?study_id=phs000424.v9.p2&object_type=variable

We therefore effectively account for their influence on gene expression, mitigating these factors' impact.

This approach represents a compromise, as it is practically infeasible to ascertain the absence of underlying health conditions in the remaining samples, even if only considering cases of “fast death”. Hence, we opted to keep all samples, independently of the cause of death of its donor, to dilute potential effects associated with individual causes of death.

(2.4) The age distribution varies across tissues which may impact the results of the study. The authors' claim that age distribution does not affect the outcomes is inconclusive. Since the study aims to provide cross-tissue analysis, it is important to note that differing age distributions across tissues can influence the overall results. To address this, the authors should conduct downsampling to different age distributions across tissues and evaluate the level of tissue-specific or common changes that remain after the distributions are made similar.

We acknowledge that variations in age distributions are evident across different tissues, with brain tissues displaying a notably pronounced disparity (green density lines in Figure R2 below).

To address this issue comprehensively, we conducted tissue-specific downsampling, by reducing the number of samples in a given age window to the minimum available sample size within all age windows for a given tissue. The histograms (density plots) of the number of samples per age window of 16 years considered in the ShARP-LM model, as well as the minimum number of samples in each age window, per tissue are illustrated in Figure R1. After performing downsampling, we computed the logFC and p-value of differential expression for each gene, per age window, and compared them (for all genes in a given age window) with those involving all samples.

Despite changes in logFC with downsampling, a considerable positive correlation is maintained (Figure R3, top panel). This suggests that the overall trends in gene expression changes persist. However, the downsampling process expectedly results in a decrease of statistical power within each age window concomitant with the decreased sample size, evident from the shift of genes from the third to the first quadrant in Figure R3, bottom panel. Consequently, we have opted for maintaining results encompassing all samples and removing the paragraph in the Discussion that asserted the absence of age distribution impact on the overall outcomes (“Indeed, we found no confounding between the distribution of samples’ ages and the trend of gene expression progression over age in any tissue.”), as we deem it inaccurate, potentially leading to misinterpretation. We have added a supplementary figure (Supplementary Figure 8, identical to Figure R3) illustrating the effect of downsampling, and the following paragraph to the manuscript’s Discussion section:

“When downsampling to ensure a balanced age distribution, a loss of statistical power is apparent but a considerable positive correlation with the original results is maintained and a substantial number of significant alterations remain so (Supplementary Figure 8).”

We acknowledge that this limitation can be addressed with the growing accumulation of human tissue transcriptomes in publicly available databases, a trend we anticipate in the near future. We are committed to promptly updating voyAGEr with any new data releases that may offer a solution to this concern.

Nonetheless, we want to underscore, as the reviewer has astutely pointed out, that while voyAGEr can facilitate cross-tissue comparisons, it must be done with caution. In this regard, we inserted the following paragraph into the Discussion:

“Due to the tissue-specific nature of the pre-processing steps (v. Read count data preprocessing in the Methods section), and given that most of the plotted gene expression distributions are centred and scaled by tissue, it is important to note that voyAGEr may not be always suited for direct comparisons between different tissues. For instance, it does not allow to directly ascertain if a gene exhibits different expression levels in different tissues or if the expression of a particular gene in one tissue changes more drastically with age than in another tissue.”

(2.5) The GTEx resource is extremely valuable, however, it comes with challenges. GTEx contains tissue samples from the same individuals across different tissues, resulting in varying degrees of overlap in sample origin across tissues as not all tissues are collected for all individuals. This could affect the similar/different patterns observed across tissues. As this tool is meant for broader use by the community, it is crucial for the authors to either rule out this possibility by conducting a cross-tissue comparison using a non-parametric model that accounts for the dependency between samples from the same individual, or to provide information on the degree of similarity between samples so that the users can keep this possibility in mind when using the tool for hypothesis generation.

We agree that the variable degrees of overlap between tissues (Figure R4) could lead to a confounding between trends in a population of common individuals and those associated with age. We therefore examined the contributions of variables 'donor,' 'tissue,' and 'age' to the overall variance in the data (Figure R5, panel A), having normalised the data collectively across all tissues. Tissue and donor contribute approximately 90% and 10% of the variance, respectively. Age exhibits minimal impact (around 1%), which may be attributed to the relative subtlety of its effects on gene expression and to the tissue specificity of ageing-associated changes. Notably, removing the 'donor' variable does not transfer this variance to 'age', suggesting a limited confounding between these variables (see Figure R5, panel B).

We also specifically examined the pairs of tissues exhibiting the lowest (Brain Amygdala / Small Intestine), median (Pancreas / Heart Left Ventricle), and highest (Kidney Cortex / Muscle Skeletal) percentages of shared donors. We identified and selectively removed samples from shared donors while maintaining the original sample size imbalance between tissues. Subsequently, we calculated each gene’s mean expression within each age window from the ShARP-LM pipeline, followed by each gene’s Pearson’s correlation of expression between tissue pairs. The resulting coefficients, both with and without the removal of common donors, were compared in scatter plots (Figure R6, left plots). As this process inherently involves downsampling, which may impact results (v. comment 2.4), we performed additional downsampling by randomly removing samples from both tissues according to the proportions defined for the removal of common donors (Figure R6, right plots).

In the chosen scenarios, we note a similar impact between the targeted removal of common donors and random downsampling. Nevertheless, the effects of removing samples may vary according to the absolute number of remaining samples. Consequently, singling out individual cases may not provide conclusive insights. To systematically address this, we represented all tissue pairs in a heatmap, colour-coded based on whether the removal of common donors is more impactful (red) or less impactful (blue) than random downsampling (Figure R7). The values depicted in the heatmap, denoted as the Impact of Common Donors (ICD), are computed for each tissue pair. This calculation involves several steps: first, we determined the absolute difference in Pearson’s correlation for each gene’s mean expression within each age window from the ShARP-LM pipeline, between the original data and the subset of data without common donors (DiffWoCD) or with random downsampling (DiffRD). Subsequently, the medians of DiffWoCD and DiffRD are computed, and the difference between these median values provides the ICD for each tissue pair. Due to the unidirectional nature of correlation (i.e., the results for tissue 1 vs tissue 2 mirror those for tissue 2 vs tissue 1), the resulting matrix is triangular in form.

We have added a supplementary figure (Supplementary Figure 4, a composition of Figures R4-R7, together with a scatterplot relating the values of heatmaps R4 and R7) that aims to provide guidance to users when interpreting specific tissue pairs, acknowledging inherent limitations (refer to comment 2.4). We have also inserted the following paragraph into the manuscript’s Discussion section:

“Furthermore, we must emphasise that the majority of GTEx donors contributed samples to multiple tissues (Supplementary Figure 4A), potentially introducing biases and confounders when comparing gene expression patterns between tissues. Our analyses of variance (Supplementary Figure 4B) and downsampling to control for common donors (Supplementary Figures 4C-E) suggest very limited global confounding between the impacts of donor and age on gene expression and that any potential cross-tissue bias not to depend much on the proportion of common donors (Supplementary Figure 4E). However, this effect must be taken into account when comparing specific pairs of tissues (e.g., Colon – Transverse and Whole Blood, Supplementary Figure 4D).”

(2.6) The authors aimed to create an open-source and ever-evolving resource that could be adapted and improved with new functionality. However, this goal was only partially achieved. Although the code for the web app is open source, crucial components such as the statistical tests or the linear model are not included in the repository, limiting the tool's customizability and adaptability.

We greatly appreciate the reviewer’s concern and share their commitment to maintaining the principles of openness, reproducibility, and adaptability for voyAGEr. voyAGEr was primarily designed as a visualisation tool, displaying pre-processed results, and indeed only the code for the Shiny app itself was accessible through the project's GitHub repository.

To address this shortcoming, we have made the entire data preprocessing script publicly available in the GitHub repository of voyAGEr. This script encompasses, among others, filtration, normalisation, batch effect correction, the ShARP-LM pipeline and statistical tests employed, and module definition. Moreover, the web app itself offers functionality to export relevant plots and tables.

(2.7) Furthermore, the authors' choice of visualization platform (R shiny) may not be the best fit for extensibility and open-source collaboration, as it lacks modularity. A more suitable alternative could be production-oriented platforms such as Flask or FastAPI.

We appreciate this thoughtful concern. The decision to use Shiny was primarily driven by our data having already been prepared in the R environment during pre-processing steps. Consequently, and as the web app serves the purpose of visualisation only (and not data processing), Shiny is as a natural and convenient extension of our scripts, enabling data visualisation seamlessly.

We acknowledge that Shiny may lack the modularity required for optimal open-source collaboration. While we recognise the merits of alternative platforms like Flask or FastAPI, we decided to keep Shiny because the current iteration of voyAGEr offers significant value to the community. Transitioning to a different platform would be a time-consuming endeavour, that would postpone the release of such resource.

However, the reviewer’s feedback regarding modularity and open-source collaboration is duly noted and highly valuable. We will certainly take it into account when developing new web applications within our laboratory.

(2.8) To facilitate collaboration and improve the tool's adaptability, data resulting from the preprocessing pipeline should be made publicly available. This would make it easier for others to contribute and extend the tool's functionality, ultimately enhancing its value for the scientific community.

As outlined in point 2.6 of this rebuttal letter, certain metadata used in our analysis are subject to restricted access. To address this, we have taken several measures to foster transparency and reproducibility of our analyses. First, we have made the scripts for data pre-processing publicly available, along with a comprehensive explanation of our methodology within the main manuscript. This empowers users to replicate our analyses and provides a foundation for those interested in contributing to the tool's development. Furthermore, we have created new issues on voyAGEr’s GitHub repository, outlining novel features and improvements we envision for the application in the future. We actively encourage users to engage with this section.

(2.9) It is unfortunate that the manuscript has no line numbers, which makes pointing out language issues or typos cumbersome. Below are some minor typos present in the current version mostly due to inconsistent usage of British vs US English, and the authors would be advised to do a thorough proofreading for the final submission.

Page 12: Inconsistent spelling of "analyzed" and "analysed". Should be "analyzed", since US English is used throughout the rest of the paper.

Page 14: "randomised"

Page 15: "emphasise"

We apologise for it and include line numbers in the revised version. We have opted for British English and corrected the manuscript accordingly.

(2.10) Some figures in the supplemental material have a low resolution (e.g. S. Fig 5). Especially figures that are not based on screenshots would ideally be of a higher resolution.

As voyAGEr is designed as a web application for visualisation, it is inherent that some screenshots of the final resource may have lower resolutions. In response to this concern, we re-generated the figures in this manuscript with a resolution that maintains clarity and readability. We also recreated figures not derived from screenshots, further improving their resolution.

We saved all figures in PDF format and are sending them together with this letter and the revised manuscript, to address any potential issues related to low-resolution figures that may occur during the export of the Word document.

<(2.11) In Fig. 1 in the bottom row the sex labels are hard to see.

We have adapted the figure to address this concern.

(2.12) Math symbols and equations are not well formatted. For example, the GE equation on p. 13, or Oiij equation should be properly typeset. Also, the Oiij notation might be confusing, I believe the authors meant to use a capital "I", i.e. OI_ij.

We have incorporated these recommendations into the revised manuscript.

(2.13) The Readme file in the git repo is very short. It would be helpful to have build and run instructions.

We have updated the README file in the GitHub repository, which now contains, among other features, instructions for launching the Shiny app and building the associated Docker image. Additionally, a simple tutorial has also been included to assist users in navigating through voyAGEr's functionalities.

(2.14> "Module" tab's UI inconsistent to other tabs (i.e. "Gene" and "Tissue"), since it contains an "About" page. Adding the "About" page in the actual "Module" page might make the UI clearer.

We believed that the Modules section, due to its distinct methodology, would benefit from an additional tab explaining its underlying rationale. We relate to the reviewer’s concern regarding the use of tabs throughout the application and made changes to the app in order to ensure consistency.

(2.15) I would suggest changing the type of the article to "Tools and Resources".

We agree and followed the reviewer’s suggestion.

Reviewer 3

(3.1) In the gene-centric analyses section of the result, to improve this manuscript and database, linear regression tests accounting for the entire range of age should be added. The authors' algorithm, ShARP-LM, tests locally within a 16-year window which makes it has lower power than the linear regression test with the whole ages. I suspect that the power reduction is strongly affected in the younger age range since a larger number of GTEx donors are enriched in old age. By adding the results from the lm tests, readers would gain more insight and evidence into how significantly their interest genes change with age.

We are grateful for the reviewer's thoughtful and pertinent recommendation and have thus conducted linear regression tests covering the entire age range. The outcomes of these tests have been integrated into the web application, denoted by a dotted orange line on the 'Gene Expression Alterations Over Age' plots. Additionally, a summary of statistics of overall changes, encompassing pvalues, t-statistics, and logFC per year, has been included below the plot title. We have also updated the manuscript to include such changes (v. Methods, Gene-centric visualisation of tissue-specific expression changes across age):

“We also applied a linear model across the entire age range, thereby providing users with more insight and supporting evidence into how a specific gene changes with age. For visualisation purposes, we incorporated a dashed orange line, with the logFC per year for the Age effect as slope, in the respective scatter plots (Figure 3B c). We depict the Sex effect therein by prominent dots on the average samples, with pink and blue denoting females and males, respectively.”

Concerning the observation about the potential reduction in statistical power due to the limited number of samples in younger ages, we acknowledge its validity. Indeed, we have addressed this issue in the manuscript's Discussion (v. Supplementary Figure 6).

(3.1) In line with the ShARP-LM test results, it is not clear which criterion was used to define the significant genes and the following enrichment analyses. I assume that the criterion is P < 0.05, but it should be clearly noted. Additionally, the authors should apply adjusted p-values for multiple-test correction. The ideal criterion is an adjusted P < 0.05. However, if none or only a handful of genes were found to be significant, the authors could relax the criteria, such as using a regular P < 0.01 or 0.05.

We apologise for any confusion regarding the terminology "significant genes." Our choice to use nonadjusted p-values for determining the significance of gene expression changes with Age, Sex, and their interaction was deliberate, and we would like to clarify our reasoning:

(1) In the "Gene" tab of the application, individual genes are examined. When users inquire about a specific gene, multiple-testing correction of the p-value does not apply.

(2) In the "Tissue" tab, using adjusted p-values and a threshold of 0.05 yielded very few differentially expressed genes, limiting the utility of Peaks. Our objective therein is not to assess the significance of alterations in individual genes but to provide a metric for global alterations within a tissue. We then determine significance based on the False Discovery Rate (FDR), using the p-values as a nominal metric of gene expression alterations.

To avoid using the concept of “differential expression”, commonly linked to significance, we now refer to 'altered genes' in both the manuscript and the app. For clarity and to align with voyAGEr's role as a hypothesis-generation tool, we define 'altered genes' as those with non-adjusted p-values < 0.01 or < 0.05, as discriminated in the Methods section.

(3.3) In the gene-centric analyses section, authors should provide a full list of donor conditions and a summary table of conditions as supplementary.

We appreciate the suggestion and we have now included a reference that directs readers to those data, alternatively to including this information as an additional supplementary table. We would like to emphasise that the web app includes information on donor conditions we hypothesise to affect gene expression.

3.4) The tissue-specific assessment section has poor sub-titles. Every title has to contain information.

We agree and revised the sub-titles to more accurately reflect the information conveyed in each corresponding section.

(3.5) I have an issue understanding the meaning of NES from GSEA in the tissue-specific assessment section. The authors performed GSEA for the DEGs against the background genes ordered by tstatistics (from positive to negative) calculated from the linear model. I understand the p-value was two-tailed, which means that both positive and negative NES are meaningful as they represent up-regulated expression direction (positive coefficient) and down-regulated expression direction (negative coefficient) with age, respectively, within a window. However, in the GSEA section of Methods, authors were not fully elaborate on this directionality but stated, "The NES for each pathway was used in subsequent analyses as a metric of its over- or downrepresentation in the Peak". The authors should clearly elaborate on how to interpret the NES from their results.

We added the following paragraph to the manuscript’s Methods section, in order to clarify the NES’ directionality:

“We extracted the GSEA normalised enrichment score (NES), which represents the degree to which a certain gene set is overrepresented at the extreme ends of the ranked list of genes. A positive NES corresponds to the gene set’s overrepresentation amongst up-regulated genes within the age window, whereas a negative NES signifies its overrepresentation amongst down-regulated genes. The NES for each pathway was used in subsequent analyses as a metric of its up- or down-regulation in the Peak.”

(3.6) In the Modules of co-expressed genes section, the authors did not explain how or why they selected the four tissues: brain, skeletal muscle, heart (left ventricle), and whole blood. This should be elaborated on.

We apologise for not providing a detailed explanation for this selection. As the ‘Modules of coexpressed genes’ section was primarily intended as a proof of concept, we opted to include tissues for which we had a substantial number of samples available and availability of comprehensive cell type signatures, those being the tissues that met such criteria. Nonetheless, as the diversity of cell type signatures increases (e.g., through the increasing availability of scRNA-seq datasets), we plan to encompass a wider range of tissues in the near future. However, as this task is time-demanding and in order to avoid a substantial delay in the release of voyAGEr, we opted to approach this issue in the next version of the App and included a dedicated issue in the projects’ GitHub repository so that users can share their preferences of the next tissues to include.

We also added a brief sentence in this regard to the Methods section of the manuscript:

“The four tissues (Brain - Cortex, Muscle - Skeletal, Heart - Left Ventricle, and Whole Blood) covered by the Module section of voyAGEr were selected due to their relatively high sample sizes and availability of comprehensive cell type signatures. The increasing availability of human tissue scRNA-seq datasets (e.g., through the Human Cell Atlas) will allow future updates of voyAGEr to encompass a wider range of tissues.”

(3.7) In the modules of the co-expressed genes section, the authors did not provide an explanation of the "diseases-manual" sub-tab of the "Pathway" tab of the voyAGEr tool. It would be helpful for readers to understand how the candidate disease list was prepared and what the results represent.

We greatly appreciate the reviewer's feedback, and in response, we have restructured the 'Modules of co-expressed genes' method section to provide a more comprehensive explanation of the 'diseases' sub-section. To clarify, we obtained a curated set of diseases and their associated genes from DisGeNET v.7.0. We assessed the enrichment of modules in relation to these diseases through two methods: a manual approach utilising Fisher’s tests (i.e. comparing the genes of a given module with the genes associated with a given disease) and another through use of the disgenet2r package, employing the function disease_enrichment. Significance of these enrichments were determined by adjusting p-values using the Benjamini-Hochberg correction.

(3.8) Most figures have low resolutions, and their fonts are too small to read.

As already mentioned in issue 2.10, we have recreated all of the images with better resolution to enhance legibility. We also exported such figures in PDF, which we attach to this revision.

(3.9) Authors used GTEx V7, which is not latest version. Although researchers have developed a huge amount of pipelines and tools for their research, most of them were neglected without a single update. I am sure many users, including myself, would appreciate it if the authors kept updating the database with GTEx V8 for the future version of the database.

We express our gratitude to the reviewer for their valuable suggestion, and, as already explained in issue 2.1, we have incorporated GTEx V8 into voyAGEr.

(3.10) I would like to have an option for downloading the results as a whole for gene, tissue, and coexpressed genes. This would be a great option for secondary analysis by users.

The implementation of such feature would be a time-demanding endeavour that would delay the release of voyAGEr, and we therefore chose not to perform it for this version. However, we agree that it would be a good resource for secondary analyses and acknowledge the possibility of adding this feature in the future. For now, voyAGEr allows the user to download all plots and corresponding data.

(3.11) How the orders of tissues in the heatmaps (both gene and tissue section) were determined? Did the authors apply hierarchical clustering? If not, I would recommend the authors perform the hierarchical clustering and add it to display the heatmap display.

We apologise for the oversight in explaining the process behind determining the order of tissues. To clarify, we employed hierarchical clustering to establish the tissue order for visualisation within the app. Although the reviewer suggested adding a dendrogram to illustrate this clustering, we decided against it. The reason for such is that including a dendrogram, while informative, is not essential for the app's primary purpose.

(3.12) I understand that this is a vast amount of work, but I hope that the authors can expand the coexpressed module analysis to include other tissues in the future version of the database.

Knowing what co-expressed genes in line with aging are and their pathway and disease enrichments across tissues would be highly informative, and I'm sure many users, including myself, would greatly appreciate it.

We express our gratitude to the reviewer for the valuable suggestion and for acknowledging the extensive effort required to incorporate new tissues into the module section. We completely agree that understanding co-expressed genes across the aging process is of significant value, and we are committed to the ongoing inclusion of additional tissues. As already stated in issue 3.6, comprehensive list of tissues slated for integration in future voyAGEr versions is readily available on voyAGEr’s GitHub repository.

Density plots (“smoothed” histograms) of the distribution of numbers of samples per moving age window for the ShARP-LM pipeline, categorised by tissue.

The numerical value within each rectangle represents the minimum number of samples observed across all age windows for that particular tissue.

Density lines (“smoothed” histograms) of the distribution of the age of donors per tissue.

As depicted in the chart, there are more samples for older ages, particularly of brain tissues.

Effect of downsampling in ShARP-LM results.

A – Per tissue violin plots of gene-wide distributions of Pearson’s correlation coefficients between original and downsampled logFC values for the Age variable across age windows, with tissues coloured by and ordered by increasing percentage of downsampling-associated reduction in the number of samples. B – Density scatter plots of comparison of associated original and downsampled p-values for each tissue, coloured by the downsampling percentage in each age window, highlighting the low range of p-values (from 0 to 0.1). Despite changes in logFC with downsampling, a considerable correlation in significance is maintained, although downsampling naturally results in a loss of statistical power, evident by the shift of points towards the first quadrant (dashed lines: p-value = 0.05).

Heatmap depicting the percentage of common donors between pairs of tissues.

A given square illustrates the percentage of all samples of tissue in the x axis (Tissue 1) that is in common with the tissue in the y axis (Tissue 2)

Assessment of the relative contributions of different sources to the dataset’s variance.

A - tissue accounts for approximately 90% of the total variance, while donor contributes around 10%; age has a minimal impact (1%), likely due to the relative subtlety of its effects on gene expression and to the tissue specificity of ageing dynamics. B - Removal of the donor variable does not transfer variance to age, suggesting limited confounding between the two variables.

Impact of the relative proportion of common donors on gene expression correlation between tissue pairs.

Panels A, B, and C showcase the tissue pairs with the highest (Muscle Skeletal / Kidney Cortex), median (Pancreas / Heart Left Ventricle), and lowest (Small Intestine / Brain Amygdala) percentages of common donors, respectively. The left panels illustrate gene-bygene Pearson’s correlations of gene expression between the two tissues, comparing the scenarios with (x-axis) and without (yaxis) the removal of common donors. The ri ght panels depict the same comparisons, but with random downsampling (y-axis) in both tissues based on the proportions defined for common donor removal. The depicted examples show that the outcomes are comparable when removing common donors or employing random downsampling.

Comparison of the impacts of removing common donor samples and random downsampling across tissue pairs.

The heatmap is coloured based on whether the removal of common donors has a greater (red) or lesser impact (blue) than random downsampling. The values depicted in the heatmap, denoted as the Impact of Common Donors (ICD), are computed for each tissue pair. This calculation involves several steps: first, by determining the absolute difference in Pearson’s correlation for each gene’s mean expression within each age window from the ShARP-LM pipeline, between the original data and the subset of data without common donors (DiffWoCD) or with random downsampling (DiffRD). Subsequently, the medians of DiffWoCD and DiffRD are computed, and the difference between these median values provides the ICD for each tissue pair. Due to the unidirectional nature of correlation (i.e., the results for tissue 1 vs tissue 2 mirror those for tissue 2 vs tissue 1), the resulting matrix is triangular in form. Grey tiles denote NA values, i.e., where the tissue-tissue comparison does not have a meaning, namely self-self and between sex-specific tissues. Top right insert: density line (“smoothed” histogram) of all ICD values.

Version 2

Abstract

Abstract

We herein introduce voyAGEr, an online graphical interface to explore age-related gene expression alterations in 49 human tissues. voyAGEr offers a visualisation and statistical toolkit for the finding and functional exploration of sex– and tissue-specific transcriptomic changes with age. In its conception, we developed a novel bioinformatics pipeline leveraging RNA sequencing data, from the GTEx project, encompassing more than 900 individuals. voyAGEr reveals transcriptomic signatures of the known asynchronous ageing between tissues, allowing the observation of tissue-specific age-periods of major transcriptional changes, associated with alterations in different biological pathways, cellular composition, and disease conditions.

Notably, voyAGEr was created to assist researchers with no expertise in bioinformatics, providing a supportive framework for elaborating, testing and refining their hypotheses on the molecular nature of human ageing and its association with pathologies, thereby also aiding in the discovery of novel therapeutic targets. voyAGEr is freely available at https://compbio.imm.medicina.ulisboa.pt/app/voyAGEr.

Introduction

The ageing-associated progressive loss of proper tissue homeostasis maintenance makes age a prevalent risk factor for many human pathologies, including cancer, neurodegenerative and cardiovascular diseases 1–3. A better comprehension of the molecular mechanisms of human ageing is thus required for the development and effective application of therapies targeting its associated morbidities.

Dynamic transcriptional alterations accompany most physiological processes occurring in human tissues 4. Transcriptomic analyses of tissue samples can thus provide snapshots of cellular states therein and insights into how their modifications over time impact tissue physiology. A small proportion of transcripts has indeed been shown to vary with age in tissue– 5 and sex-specific 6–11 manners. Such variations reflect dysregulations of gene expression that underlie cellular disfunctions 5.

Many studies analysed the age-related changes in gene expression in rodent tissues 12–18, emphasising the role in ageing of genes related to inflammatory responses, cell cycle or the electron transport chain. However, while it is possible to monitor the modifications in gene expression over time in those species by sequencing transcriptomes of organs of littermates at different ages, as a close surrogate of longitudinality, such studies cannot be conducted in humans for ethical reasons. Indeed, most studies aimed at profiling ageing-related gene expression changes in human tissues focused on a single tissue (e.g. muscle 19–21, kidney 22, brain 23–27, skin 28,29, blood 30,31, liver 32, retina 33) and/or are limited to a comparison between young and old individuals 20,21,28,32,33, failing to fully capture the changes of the tissue-specific gene expression landscape throughout ageing 5. A few studies were nonetheless led on more than one tissue in humans, from post-mortem samples 34,35 and biopsies 36,37, and in mice 18,36 and rats 38. The age-related transcriptional profiles derived therein, either from regression 18,34,35,37 or comparison between age groups 36,38, highlight an asynchronous ageing of tissues (discussed in 39), with some of them more affected by age-related gene expression changes associated with biological mechanisms known to be impacted by ageing such as mitochondrial activity or metabolic homeostasis. In particular, tissue-specific periods of major transcriptional changes in the fifth and eighth decades of human lifespan have been revealed 34, reflecting the so-called digital ageing 39, consistent with what is observed in mice 17,18. Furthermore, despite outlining the tissue specificity of the transcriptomic signatures of human ageing, some synchronisation was found between tissues like lung, heart and whole blood, which exhibit a co-ageing pattern 35. Nevertheless, as each study followed its own specific procedures, from sample collection to data processing, results from these analyses are hard to compare with one another.

Processed data from those studies have not been made easily accessible and interpretable to researchers lacking computational proficiency but aiming to use them to test their novel hypotheses. To fill this void, we have developed voyAGEr, a web application providing flexible visualisation of comprehensive functional analyses of gene expression alterations occurring in 49 human tissues with age in each biological sex. We leverage the large RNA-seq dataset from the Genotype-Tissue Expression (GTEx) project 40, encompassing tissue samples from hundreds of donors aged from 20 to 70 years, with a pipeline for gene expression profiling with an optimised temporal resolution. voyAGEr allows to investigate ageing from two perspectives: (i) gene-centric – how each gene’s tissue-specific expression progresses with age; and (ii) tissue-centric – how tissue-specific transcriptomes change with age. Additionally, voyAGEr enables the examination of modules of co-expressed genes altered with age in 4 tissues (brain cortex, skeletal muscle, left ventricle of the heart, whole blood), namely their enrichment in specific cell types, biological pathways, and association with diseases. We therefore expect voyAGEr to become a valuable support tool for researchers aiming to uncover the molecular mechanisms underlying human ageing.

Moreover, being open-source, voyAGEr can be adapted by fellow developers to be used with alternative datasets (e.g. from other species) or to incorporate other specific functionalities.

voyAGEr is freely available at https://compbio.imm.medicina.ulisboa.pt/app/voyAGEr.

Results

voyAGEr’s interactive exploration of tissue-specific gene expression landscapes in ageing is based on sequential fitting of linear models (v. Materials and Methods) to estimate, for each gene in each tissue:

the Age effect, i.e., how the age-associated changes in gene expression evolve with age itself;

the Sex effect, i.e., how the differences in gene expression between sexes evolve with age;

the Age&Sex interaction effect, i.e., how the differences between sexes of age-associated changes in gene expression evolve with age.

We named our approach Shifting Age Range Pipeline for Linear Modelling (ShARP-LM). Briefly, this method consists in performing differential gene expression (with gene expression as a function of the donors’ Age, Sex and Age&Sex interaction) in moving age windows spanning 16 years. By considering the percentage of genes altered in each age range, we can highlight age periods of major tissue-specific transcriptomic alterations (Figure 1).

voyAGEr profiles tissue-specific age-associated changes in gene expression and their differences between sexes.

For each of the 49 human tissues in GTEx, gene expression was linearly modelled in windows spanning 16 years centred in consecutive years of age, to estimate the effects thereon of Age, Sex and the interaction between them, i.e., how the Sex effect changes with age, equivalent to how the Age effect differs between sexes (v. Materials and Methods). In each age window, the percentage of genes with expression significantly altered by of each of those effects gives their respective transcriptomic impact (upper panels). voyAGEr thereby identifies the age periods at which major gene expression changes occur in each tissue. For example, in coronary artery: major age-related transcriptional alterations are found at around 60 years of age (upper left panel), illustrated by the behaviour of COX11 (bottom left panel); major gene expression differences between males and females happen across the considered age range (upper centre panel), as illustrated by CD99 (bottom centre panel); major differences between sexes in age-related gene expression alterations happen across the considered age range (upper right panel), as illustrated by AKT2 (bottom right panel). Solid loess lines in the bottom panels (green for all donors, pink for females, blue for males). Gene expression (GE) in log2 of counts per million (logCPM).

Gene-centric analyses of human tissue-specific expression changes across age

The progression of tissue-specific expression of a particular gene across age can be examined in voyAGEr’s Gene tab. By entering its HGNC symbol in the Gene selector, the user has access to graphical summaries of the gene’s tissue-specific expression (sub-tab Profile) (Figure 2A) and significance of age-related changes in its expression due to the Age, Sex, and Age&Sex effects (sub-tab Alteration) (Figure 2B) across age. Results can be displayed in a heatmap for all tissues or in a scatter plot for a chosen individual tissue (Figure 2C). When the gene is studied in a single tissue, the user can graphically and statistically profile the association of the donors’ sex and reported conditions (e.g., history of heart attack or pneumonia) with the gene’s expression profile. A table summarizing the donors’ metadata is also shown (Figure 2C). The user can interactively select donors of interest on the scatter plot and further examine their information in the automatically subsetted table.

Gene-centric analyses of expression alterations across age

A – Heatmap of MKI67 expression across tissues over age. B – Heatmap of significance of Age-associated MKI67 expression alterations over age. p-values are for the moderated t-statistics of differential gene expression associated with the Age effect (v. the ShARP-LM approach in Materials and Methods). Notably, transcriptional changes are observed in lung (mid 20’s, early 30’s and after 55). C – voyAGEr’s Gene tab interface. MKI67 expression in lung is inspected. Donors’ information is shown in a table and the scatter plot can be interactively adjusted according to the donors’ condition of interest (Supplementary Figure 1B).

An example of a process whose molecular mechanisms are of particular interest to researchers in the ageing field is cellular senescence. Senescence is a stress-induced cell cycle arrest limiting proliferation of potentially oncogenic cells but progressively creating an inflammatory environment in tissues as they age 41,42. CDKN2A, that encodes cell cycle regulatory protein p16INK4A known to accumulate in senescent cells 43,44, has its expression increased with age in the vast majority of tissues profiled (Supplementary Figure 1A).

Similarly, reduced levels of proliferation markers, such as PCNA 45 and MKI67 46, can be studied as putative markers of ageing of certain tissues. These genes have their expression altered with age in the lung and display a similar expression profile (decreasing from 25 to 30 years old (y.o.), constant between 35 and 50 y.o. and decreasing in older ages) (Figure 2C). However, these trends appear to vary according to the donor’s history of non-metastatic cancer (Supplementary Figure 1B), illustrating voyAGEr’s use in helping to find associations between gene expression and age-related diseases.

On a different note, sex biases have been reported in the expression of SALL1 and DDX43 in adipose tissue and lung, respectively 10. voyAGEr allows to not only recapitulate those observations but also assess the temporal window where these changes occur (Supplementary Figure 2).

Tissue-specific assessment of gene expression changes across age

Peaks of gene expression alterations

The landscape of global tissue-specific gene expression alteration across age can be examined in voyAGEr’s Tissue main tab. A heatmap displaying, for all tissues, the statistical significance over age (v. Materials and Methods) of the proportion of genes altered with Age, Sex or Age&Sex (depending on the user’s interest) is initially shown (Figure 3A), illustrating the aforementioned asynchronous ageing of tissues observed for humans and rodents 18,32–36,39.

Tissue-specific assessment of gene expression changes across age

A – Heatmap of significance (FDR based on random permutations of age, v. Materials and Methods) over age of the proportion of genes with expression significantly altered with Age in the 49 analysed tissues. B – Exploration of gene expression changes across age in Subcutaneous Adipose tissue: (a) Percentage of genes with significantly altered expression with Age over age. Two main peaks of transcriptional changes are noteworthy, a major one at the late 20’s and a minor one after 45; (b) Clicking on the dot of a specific age (29.57 y.o. in plot a) gives access to the list of the most altered genes at that age, ordered by statistical significance of expression changes (p-value of moderated t-statistic). (c) Plot of expression of the chosen top gene in the table in b across age (bottom) in parallel with the significance of its expression alterations with Age. The expression of LMO3 significantly increases at around 30 years old, concomitantly with the first peak of transcriptomic changes with Age.

The user can then spot the age periods with the most significant gene expression alterations in a selected tissue (Figure 3B a), and identify the associated altered genes (Figure 3B b).

The user can also plot the expression of a given gene of interest across age together with the significance of its expression modification with Age, Sex or Age&Sex (Figure 3B c).

An example of voyAGEr’s capabilities is illustrated in Supplementary Figure 3, showing substantial transcriptomic alterations in the uterus from the late forties to the early fifties, overlapping with the age distribution of menopausal onset47, which could explain the observed molecular modifications.

It is also possible to visualise tissues with more than one period of transcriptomic changes, and to individually inspect these periods. As an example, the subcutaneous adipose tissue appears to go through two main periods of transcriptional changes with age: a major one at the late 20’s (∼13 % of altered genes), and a minor one after 45 years (∼5% of altered genes) (Figure 3B a). The most altered genes in this first peak appear to have their expression modified only at this precise age period (e.g. PRELID1, RUNX1T1, TUBB4B, FGFRL1 and MALSU1). Similarly, mitochondrial genes (e.g. MT-CYB, MT-ND4, MT-ATP6, MT-ND2) (Figure 3B b) appear to be the most altered genes in the second peak (Figure 3B c). This particularity suggests that different sets of genes drive the periods of major transcriptional changes, which begs assessing if they reflect the activation of distinct biological processes.

Gene Set Enrichment

The user can explore the biological functions of the set of genes underlying each peak of transcriptomic changes by assessing their enrichment in curated pathways from the Reactome database 48. voyAGEr performs Gene Set Enrichment Analysis (GSEA) 49 and the user can visualise heatmaps displaying the evolution over age of the resulting normalised enrichment score (NES49, reflecting the degree to which a pathway is over– or under-represented in a subset of genes) for a given tissue, effect (Age, Sex or Age&Sex) and Reactome pathway (all, or user-selected) (Figure 4A a). To reduce redundancy and facilitate the understanding of their biological relevance, we clustered those pathways into families that also include Kyoto Encyclopedia of Genes and Genomes (KEGG) pathways 50 and Gene Ontology (GO) Biological Processes of level 3 51. Briefly, we clustered gene sets from the 3 sources based on the overlap of their genes (v. Materials and Methods), thereby creating families of highly functionally related pathways. Taking advantage of the complementary and distinct terminology in Reactome, KEGG and GO, the user can interpret each family’s broad biological function by looking at the word cloud of its most prevalent terms, and browsing the list of its associated pathways (Figure 4A b). For example, although most pathways enriched in the two aforementioned peaks of altered genes in subcutaneous adipose tissue were different, there is an overlap of pathways related to metabolism, including various mitochondrial processes (Figure 4A). This highlights the importance of integrating individual gene data with pathway enrichment analysis to garner more comprehensive insights on the biological relevance of those changes.

Tissue-specific assessment of pathway expression changes across age in the human Subcutaneous Adipose tissue

A – Heatmap depicting the normalised enrichment scores (NES) of Reactome pathways associated with specific tissues and effects. Pathways are classified into 10 families (a), which can be characterised by their frequently occurring terms (b), providing insights into their biological functions. Only pathways significantly linked to gene expression changes in at least one age window (FDR ≤ 0.01) are displayed. Black squares indicate the two age periods with prominent transcriptional changes, while yellow squares denote pathways common to both peaks, primarily belonging to family 2. Word cloud analysis (b) reveals that family 2 pathways are mainly related to metabolism. B – Enrichment of a user-provided gene set, given by the significance of Fisher’s tests, in genes altered with Age throughout ageing (based on a user-defined p-value threshold). Here, the given gene set is composed of genes from Senequest 42 whose link with senescence is supported by at least 4 sources. In this case, there are no significant peaks.

The peaks of transcriptomic changes can also be examined for enrichment in a user-provided gene set (Figure 4B). As expression of senescence-related CDKN2A is increased in the subcutaneous adipose tissue with age (Supplementary Figure 1A), we hypothesised that other senescence-associated genes may be augmented too. Thus we used that voyAGEr functionality, using the Senequest 42 geneset (supported by at least 4 sources) to test it, observing no significant alterations (Figure 4B).

Modules of co-expressed genes

voyAGEr also allows functional analyses of modules of co-expressed genes (i.e. genes with highly correlated expression across samples, defined by weighted correlation network analysis 52). Genes in the same module are likely to be co-regulated and share biological functions or associations with phenotypical or pathological traits 53. Those modules may also act as markers of core transcriptional features of cellular activity and identity 54.

Concretely, voyAGEr enables the user to visualise how the expression of modules of genes that are associated with a specific cell type, biological pathway or disease progresses over age in a specific tissue. After selecting one tissue of interest, the user has, for each module, access to four levels of information:

Expression: its eigengene expression progression over age (Figure 5C);

Cell types: its enrichment in specific cell types, based on cell type signatures found in the literature (Figure 5C);

Pathways: its enrichment in Reactome pathways;

Diseases: its enrichment in disease markers, based on gene-disease associations from DisGeNET 55,56, calculated with both the disgenet2r package 55 and with Fisher’s tests (Figure 5B).

Tissue-specific assessment of age-associated progression of modules of co-expressed genes

A – Heatmap of eigengene expression for all the modules of co-expressed genes in the brain cortex over age. B – Heatmap of association of the modules with four selected diseases, computed with the disgenet2r package55. C – Scatter plot (above) of eigengene expression over age, in all brain cortex samples, for a selected module of 241 genes (MEblue). The eigengene expression is derived from the 1st component of the single value decomposition of its genes’ expression. This module was analysed, with Fisher’s tests, for enrichment in cell types, based on markers from the literature (Fan84 and Descartes85) and found to be associated with astrocytes, as can be observed by the TreeMap below (where each rectangle’s area and darkness are proportional to the significance of its association with a cell type and its colour linked to the markers’ source study).

By default, for each tissue, results are displayed in the form of heatmaps of expression (centred and scaled) illustrating how all modules evolve with age (Figure 5A). The user also has the possibility to select a module of interest and see its eigengene progression over age in a scatter plot (Figure 5C), lists of its association with diseases and pathways ordered by significance, and a TreeMap for its cell type enrichments (Figure 5C). In the example of Figure 5C, the “MEblue” module, comprising 241 genes co-expressed in the brain cortex, shows significant enrichment in astrocyte markers. The apparent increase of this module’s expression with age may reflect the known age-related changes in astrocyte activation 57 and the relative weakening of neuronal activity (“MEyellow” and “MEpink” modules).

As in the Gene tab, the user can separate donors based on their sex and associated medical conditions in the scatter plot of the eigengene expression progression. On the Pathways and Diseases-Manual tabs below the main plot, the user can also visualise the contingency table from that specific disease/pathway, on the corresponding column.

Discussion

voyAGEr provides a framework to examine the progression of gene expression over age in several human tissues, serving as a valuable resource for the ageing research community. In particular, it helps to identify tissue-specific age periods of major transcriptomic alterations. The results of our analyses show the complexity of human biological ageing by stressing its tissue specificity 34 and non-linear transcriptional progression throughout lifetime, consistent with previous results from both proteomic 58 and transcriptomic 28,34 analyses. By revealing and annotating the age-specific transcriptional trends in each tissue, voyAGEr aims at assisting researchers in deciphering the cellular and molecular mechanisms associated with the age-related physiological decline across the human body.

Due to the tissue-specific nature of the pre-processing steps (v. Read count data pre-processing in the Methods section), and given that most of the plotted gene expression distributions are centred and scaled by tissue, it is important to note that voyAGEr may not be always suited for direct comparisons between different tissues. For instance, it does not allow to directly ascertain if a gene exhibits different expression levels in different tissues or if the expression of a particular gene in one tissue changes more drastically with age than in another tissue. Furthermore, we must emphasise that the majority of GTEx donors contributed samples to multiple tissues (Supplementary Figure 4A), potentially introducing biases and confounders when comparing gene expression patterns between tissues. Our analyses of variance (Supplementary Figure 4B) and downsampling to control for common donors (Supplementary Figures 4C-E) suggest very limited global confounding between the impacts of donor and age on gene expression and that any potential cross-tissue bias not to depend much on the proportion of common donors (Supplementary Figure 4E). However, this effect must be taken into account when comparing specific pairs of tissues (e.g., Colon – Transverse and Whole Blood, Supplementary Figure 4D).

Additionally, voyAGEr allows to scrutinise and visually display the tissue-specific differences in gene expression between biological sexes across age. Biological sex is an important factor in the prevalence of ageing-associated diseases, as well as their age of onset, progression 6,7,59 and sex-related differences in gene expression8–11. By profiling the age distribution of such differences, voyAGEr can lead to a better understanding of their influence in the aetiology of the sex specificities of human ageing. For instance, we were able to corroborate findings on the sex-differential transcriptome of adult humans by Gershoni et al. 9, with voyAGEr emphasising its tissue-specificity and allowing to discriminate the ages at which sex-related biases appear to be more prevalent (Supplementary Figure 5).

Nonetheless, it is essential to interpret sex chromosome-specific gene results in voyAGEr with caution. For instance, we observe elevated expression of Y-chromosome-specific DDX3Y in males, whilst its female expression (expected to be zero) is very low, in the range of what can be considered background noise (Supplementary Figure 6). Its age-related alterations exhibit a distinctive peak around the age of 40 apparently driven by subtle changes of gene expression in female samples, illustrating the need for the abovementioned caution.

One of the limitations of voyAGEr is that most GTEx tissue donors had health conditions and their frequency increased with age, preventing us from defining a class of healthy individuals and identifying age-associated transcriptomic changes that could be more confidently proposed to happen independently of any disease progression. Whilst the large sample sizes and inherent biological variability among individuals, reflected in the diversity of condition combinations, are expected to mitigate significant confounding effects, voyAGEr also allows users to evaluate how tissue-specific gene expression trends vary according to the donors’ diverse conditions (see Supplementary Figure 1B).

The development of voyAGEr was accompanied by that of a pipeline, ShARP-LM, that facilitates the holistic depiction of the transcriptional landscapes of adult human tissues throughout ageing with a yearly age resolution. We take advantage of the comprehensiveness of the transcriptome collection from human tissues from the GTEx project to make our analyses a valid surrogate of a currently undoable longitudinal study. It confers our method enough statistical robustness to mitigate the inter-individual differences and deal with the non-uniform distribution of the donors’ ages. Nevertheless, it is worth highlighting that the age distribution of donors does impact the statistical power for detecting transcriptional changes. Consequently, we are more likely to identify significant alterations (with p-value < 0.05 in our gene-centric analyses) within age ranges that are more prevalent in our sample population, often characterised by older individuals (Supplementary Figure 7). When downsampling to ensure a balanced age distribution, a loss of statistical power is apparent but a considerable positive correlation with the original results is maintained and a substantial number of significant alterations remain so (Supplementary Figure 8). This limitation is likely to be overcome by the accumulation of transcriptomes of human tissues in public databases, promising a gradual increase in accuracy and age resolution with which human transcriptomic ageing can be profiled.

Similarly, the expanding collection of single-cell transcriptomes in public databases is yielding improved gene markers for an increasing diversity of human cell types, enhancing the usefulness of leveraging bulk transcriptomes to study the impact of ageing on the cellular composition of human tissues, for which the co-expression module approach in voyAGEr provides a proof-of-concept.

Nonetheless, it is imperative to approach the module-based analysis with caution, as direct and literal interpretations may be misleading. For instance, it is not uncommon to observe an enrichment of “Rheumatoid Arthritis” in modules associated with various immune cell types in anatomical locations, such as the brain cortex, where the disease does not directly manifest. If a specific module associated with a condition like “Liver Cirrhosis” exhibits an age-related increase in the brain cortex, of course this does not mean such disease ever occurs within older brains. Nevertheless, we consider that the module-based approach can serve as a valuable resource for generating hypotheses.

Given its open-source nature, voyAGEr is envisaged to be a continually evolving resource, able to accommodate new data and expand its functionalities, namely by incorporating additional tissues into the modules section and integrating perturbagen data for inference of molecular causes underlying observed gene expression alterations and small molecules to target them for therapeutic purposes 60,61.

As an in silico approach with no experimental validation for its results, voyAGEr is meant to be a discovery tool, supporting biologists in the exploration of a large transcriptomic dataset, thereby generating, refining or preliminary testing hypotheses, laying the groundwork for subsequent experimental research. It can be an entry point for projects aiming at better understanding the tissue– and sex-specific transcriptional alterations underlying human ageing, to be followed by targeted studies focusing on the functional roles of the most promising markers identified therein in the physiology of ageing. Those marker genes can contribute to the development of more robust and cross-tissue gene signatures of ageing 62 and the expansion of age-related gene databases 63,64.

Moreover, the observed diverse and asynchronous changes in gene expression between tissues over the human adult life provides potentially relevant information for the design of accurate diagnostic tools and personalised therapies. On one hand, identifying those changes’ association with specific disorders could have a prognostic value by enabling the identification of their onset before clinical symptoms manifest 36. On the other hand, computational screening of databases of genetic and pharmacologically induced human transcriptomic changes could help to infer their molecular causes and uncover candidate drugs to delay these effects 60,61,65,66.

Methods

Development platforms

Data analysis was performed in R (version 4.1.2) and the application developed with R package Shiny 67. voyAGEr’s outputs are plots and tables, generated with R packages highcharter 68 and DT 69, respectively, that can easily be downloaded in standard formats (png, jpeg, and pdf for the plots; xls and csv for the tables).

voyAGEr was deployed using Docker Compose and ShinyProxy 2.6.1 in a Linux virtual machine (64GB RAM, 16 CPU threads and 200GB SSD) running in our institutional computing cluster.

Read count data pre-processing

The matrix with the RNA-seq read counts for each gene in each GTEx v8 sample was downloaded from the project’s data portal (https://www.gtexportal.org/) 40. From the 54 tissues available from GTEx v8, five were discarded (kidney medulla, fallopian tube, bladder, ectocervix, endocervix) due to low (<50) numbers of samples.

Read count data for each tissue were then pre-processed separately. We started by filtering out genes deemed uninformative due to their very low expression across samples: only genes with at least 1 CPM in at least 40% of the samples were kept for analysis (the number of genes analysed for each tissue can be found in Supplementary Table 1)70. Read counts for those kept genes were used to calculate normalisation factors to scale the raw library sizes, using function calcNormFactors from edgeR 70 that implements the trimmed mean of M-values 71. Read counts were subsequently normalised and log-transformed with the voom function 72 from package limma 73.

However, it is well-established that batch effects, which may stem from variations in sample treatment prior to RNA-seq library preparation, can introduce spurious gene expression differences between samples and result in confounding factors74. We therefore conducted an impartial and systematic search for potential batch effects. Firstly, we performed a principal component analysis of gene expression for each tissue, using the prcomp package.

We quantified the relation between each condition associated with every sample1 and the first two principal components, by computing Spearman correlations (for numerical conditions), t-tests (for binary categorical conditions), or analysis of variance (ANOVA) tests (for variables with more than two possible values, and, in the case of numerical variables, fewer than 15 unique values). Conditions that surpassed defined empirical thresholds (p-value < 0.05, Spearman correlation > 0.3, t-test > 10, and ANOVA > 20) were flagged as potential batch effects. Except in brain tissues, the COHORT variable (i.e., whether the samples were collected from organ donors or post mortem) appeared to be the main batch effect, with ripple effects on numerous other related conditions (Supplementary Figure 9). Moreover, SMRIN (sample’s RNA integrity number), DTHHRDY (death classification based on the 4-point Hardy Scale), and MHSMKYRS (smoke years) consistently emerged as conditions associated with the primary axes of variance. The number of genes detected in each sample, determined by the filtration step described above, was also identified as a significant contributor to the primary data variance. We therefore corrected for these five conditions, on a tissue-by-tissue basis, by adapting the removeBatchEffect function from the limma package 73. Specifically, we employed linear models to estimate the contributions to gene expression of each of those factors and subtracted such contributions from the original logCPM matrix. To ensure biological interpretability of the results, we offset the resulting values to the minimum value in the non-corrected matrix. To prevent sample loss due to missing values for the aforementioned five conditions and since the number of missing values was relatively low compared to the total number of samples, imputation was carried out using the mice package75.

The resulting matrix of logCPM-corrected values was used for all downstream analyses. As an illustrative example of the importance of batch removal, the expression of surfactant factor SFTPA2 was found to be associated with donors on a ventilator76. Without batch correction, SFTPA2 expression would have been associated with age due to higher prevalence of such cases among older individuals (Supplementary Figure 10).

ShARP-LM

To model the changes in gene expression with age, we developed the Shifting Age Range Pipeline for Linear Modelling (ShARP-LM). For each tissue, we fitted linear models to the gene expression of samples from donors with ages within windows with a range of 16 years shifted through consecutive years of age (i.e. in a sliding window with window size = 16 and step size = 1 years of age). This was the minimum age span needed to guarantee the presence of more than one sample per window, across all considered tissues. As samples at the ends of the dataset’s age range would be thereby involved in fewer linear models, we made the window size gradually increase from 11 to 16 years when starting from the “youngest” samples and decrease from 16 to 11 years when reaching the “oldest” (Supplementary Figure 11).

Function lm from limma was used to fit the following linear model for gene expression (GE): For each gene, α, β and X are the coefficients to be estimated for their respective hypothesised effects. For each sample, Age in years and Sex in binary were centred and Age&Sex interaction given by their product. The coefficients E0 and χ are thus the expression of the average sample (i.e., with average sex and average age) and the error term.

For each gene in each model (i.e., each age window in each tissue), we retrieved the t-statistics of differential expression associated with the three relevant variables and their respective p-values. We considered the average age of the samples’ donors within the age window as the representative age of the observed expression changes.

In summary, for a given tissue and variable (Age, Sex and Age&Sex), ShARP-LM yields t-statistics and p-values over age for all genes, reflecting the magnitude and significance of the changes in their expression throughout adult life.

Gene-centric visualisation of tissue-specific expression changes across age

For visualisation purposes, the trend of each gene’s expression progression over age in each tissue was derived through Local Polynomial Regression Fitting, using R function loess on logCPM values (Figures 1, 2C, 3B c). For summarizing in a heatmap a given gene’s expression across age in multiple tissues (Figure 2A) or the expression of several genes across age in a specific tissue, each gene’s regression values are centred and scaled, using R function scale.

For summarizing in a heatmap the significance of a given gene’s expression changes over age in multiple tissues (Figure 2B), cubic smoothing splines are fitted to –log10(p), with p being the t-statistic’s p-value, with R function smooth.spline.

Tissue-specific quantification of global transcriptomic alterations across age

To assess the global transcriptomic impact of each of the three modelled effects in each tissue across age, we analysed the progression over age (i.e. over consecutive linear models) of the percentage of genes whose expression is significantly altered (t-statistic’s p-value ≤ 0.01) by each effect (Figure 3B). To evaluate the significance of each percentage and assess if high percentages can be confidently associated with major transcriptomic alterations, we controlled for their false discovery rate (FDR) by randomly permuting the samples’ ages and sexes within each age window fifty thousand times and performing ShARP-LM on each randomised dataset. We were then able to associate an FDR to each percentage of altered genes by comparing it with the distribution of those randomly generated (Figure 3B a).

We also applied a linear model across the entire age range, thereby providing users with more insight and supporting evidence into how a specific gene changes with age. For visualisation purposes, we incorporated a dashed orange line, with the logFC per year for the Age effect as slope, in the respective scatter plots (Figure 3B c). We depict the Sex effect therein by prominent dots on the average samples, with pink and blue denoting females and males, respectively.

Gene Set Enrichment Analysis (GSEA)

For each Peak of significant gene expression modifications, we performed GSEA 49 on the ordered (from the most positive to the most negative) t-statistics of differential expression for the respective tissue and age, using R package fgsea 77 and the Reactome database 48. We extracted the GSEA normalised enrichment score (NES), which represents the degree to which a certain gene set is overrepresented at the extreme ends of the ranked list of genes. A positive NES corresponds to the gene set’s overrepresentation amongst up-regulated genes within the age window, whereas a negative NES signifies its overrepresentation amongst down-regulated genes. The NES for each pathway was used in subsequent analyses as a metric of its up– or down-regulation in the Peak. The resulting normalised enrichment score (NES) for each pathway was used in subsequent analyses as a metric of its over– or down-representation in the Peak.

To optimise computational efficiency and minimise redundancy in the analysed pathways, we only considered pathways containing a minimum of 15 genes and up to 500 genes, as suggested in the GSEA User Guide78. For the sake of clarity in voyAGEr’s visual representations, we only included pathways with a p-adjusted value less than or equal to 0.05, further narrowing it down to pathways within the top 1% of p-adjusted values.

Additionally, we exclusively featured pathways with at least one significant age Peak (FDR <= 0.05), as illustrated in Figure 4A.

Families of pathways

To reduce pathway redundancy and facilitate the assessment of their biological relevance in results’ interpretation, we created an unifying representation of pathways from Reactome 48, KEGG 50, and level 3 Biological Processes from GO 51, by adapting a published pathway clustering approach 79 to integrate them into families.

The approach relies on the definition of a hierarchy of pathways based on the numbers of genes they have in common. For each two pathways Pi and Pj, respectively containing sets of genes Gi and Gj, we computed their overlap index (OI) 79, defined as follows: Where |Gi| is the number of genes in set Gi and |GiνGj| is the number of genes in common between Gi and Gj. OIij = 1 would therefore indicate that Pi and Pj are identical in gene composition or that one is a subset of the other. On the contrary, OIij = 0 would mean that Pi and Pj have no genes in common. To ease the computational work, we removed from the analysis pathways that are subsets of larger pathways (i.e. each pathway whose genes are all present in another pathway).

From the OIij matrix, from which each row is a vector with the gene overlaps of pathway i with each of the pathways, we computed the matrix of Pearson’s correlation between all pathways’ overlap indexes with R function cor. That matrix was finally transformed into Euclidean distances with R function dist, allowing for pathways to be subsequently clustered with the complete linkage method with R function hclust. The final dendrogram was empirically cut in 10 clusters (Supplementary Figure 12). Pathways that were initially excluded from the computation for being subsets of others were added to the clusters of their respective parent pathways. Each daughter pathway with more than one parent was assigned to the cluster of the parent with the smallest number of genes, thereby maximizing the daughter-parent similarity. The data.table R package, for fast handling of large matrices, was used in this analysis 80.

Gene co-expression modules

Gene co-expression modules were defined with R package WGCNA52. For each considered tissue, the process began with the identification of a set of informative genes that exhibit high variability across samples (referred to as variable A), thus improving module definition. Next, we calculated the bicorrelation matrix for the expression of all selected genes with the bicor function. We then applied soft thresholding by raising all correlation values to the power of β, accentuating stronger correlations. The value of β = 12 was chosen in accordance to the WGCNA FAQs81, and after confirmation of a free-scale topology using the pickSoftThreshold function. We generated the dissimilarity matrix by subtracting the output of the TOMsimilarity function from 1. Gene co-expression modules were then defined using a static tree-cutting algorithm, implemented via the cutreeStaticColor function, requiring as parameters a minimum number of genes per module (referred to as variable B) and the tree-cutting height (variable C).

The empirical determination of parameters A to C was guided by the following principles: (i) maximizing cell type signature enrichment, (ii) minimizing the number of genes per module, and (iii) ensuring that the modules’ eigengenes exhibited age-related variability. Different combinations of these variables were exhaustively tested. To maintain biological relevance, modules consisting of non-assigned genes or genes lacking substantial supporting evidence, such as pseudogenes, were excluded.

Maximizing cell type enrichment in the modules, with a focus on known markers for specific cell types, has previously been proven successful in unveiling core transcriptional features of cell types in the human central nervous system 53. Cell type enrichment analysis relied on Fisher tests, providing odds enrichment scores and significance values (p-values). This involved comparing module genes with the signature, considering as background of genes those included in the module definition for each tissue. We prioritised modules of genes for which we obtained at least one significant result for each cell type (odds ratio (OR) > 1 and p-value < 0.05). The cell type signatures employed in this analysis were sourced from MSigDB’s C8 collection49, and in the case of Whole Blood, additionally from LM2282.

Specific variance thresholds (variable A) were employed: 0.5, 0.4, 0.35, and 0.9 for Brain – Cortex, Muscle – Skeletal, Heart – Left Ventricle, and Whole Blood, respectively. The minimum number of genes per module (variable B) was set at 15, 20, 20, and 15, respectively. Tree-cutting heights (variable C) of 0.95, 0.98, 0.99, and 0.97 were respectively used.

Each module is characterised by a set of genes and an eigengene, represented by the first principal component obtained through singular-value decomposition of the module’s gene expressions. Subsequently, voyAGEr facilitates an evaluation of cell type enrichment, as described earlier, and enrichment in biological pathways and diseases. The enrichment of modules in cell types, Reactome pathways, and diseases (extracted from DisGeNET 55,56 database version 7.0) was quantified using Fisher’s tests. For disease enrichment, the function disease_enrichment from the disgenet2r package55 was employed, utilising their curated set of diseases. The significance of these enrichments was determined through p-value/FDR adjustment using Benjamini-Hochberg correction. For visual clarity, only pathways and diseases displaying significant enrichment (p <= 0.05) in at least one module were considered.

The four tissues (Brain – Cortex, Muscle – Skeletal, Heart – Left Ventricle, and Whole Blood) covered by the Module section of voyAGEr were selected due to their relatively high sample sizes and availability of comprehensive cell type signatures. The increasing availability of human tissue scRNA-seq datasets (e.g., through the Human Cell Atlas83) will allow future updates of voyAGEr to encompass a wider range of tissues.

Data and Code availability

Processed GTEx v8 RNA-seq data (read count tables) were downloaded from the project’s data portal (https://www.gtexportal.org/). Donor metadata were obtained from dbGaP – database of Genotypes and Phenotypes (Accession phs000424.v9.p2 project ID 13661). voyAGEr’s output tables can be directly downloaded in standard xls and csv formats.

The complete source code for voyAGEr, including pre-processing and Shiny app, can be accessed on GitHub at the following repository: https://github.com/DiseaseTranscriptomicsLab/voyAGEr.